### 2004-03-12 doc/paper/skysubtraction.tex
Skysubtraction, not much here yet

The object extraction lists 5 steps in header:

\begin{enumerate}
\item{Locate bright fibers (aka whopping fibers)}
\item{Assess sky fiber levels}
\item{Commence iterative Optimal extraction }
\item{Shift wavelength solution to match sky lines (describe in wavelength solution)}
\item{Sky subtract}

\end{enumerate}

Note: Flux correction and Telluric correction have been moved (a long time ago).

\subsection{Sky Subtraction}

A global sky model is fit using all available fiber spectra assigned as 
"SKY" by target selection.  These fibers are set on blank fields, and are 
used to estimate the sky level as a function of wavelength and CCD position 
in individual exposures.   Each plate is nominally assigned 32 of its 640 
fibers as "SKY", with the focal plane positions chosen to allow half of the 
sky fibers to reach each spectrograph: 1 fiber per bundle of 20 is assigned to 
"SKY".

The sky subtraction is done after the data have been extracted from 
two-dimensional counts per CCD pixel to one-dimensional "corrected" electron 
counts as a function of vacuum heliocentric wavelengths and fiber number.
The flat-field "correction" removes differences due to surface brightness 
changes due to varying spectral dispersion in different fibers.

A single subroutine is called three consecutive times to arrive at the final 
model description for the sky background.  The first call uses all available sky
fibers that were not masked as an entire fiber (e.g. by the fibermask bits).
If only one good skyfiber is found, the routine logs an "ABORT" and returns.
A median count level is summarized for the selected sky spectra, and the 5% 
and 95% levels are reported for each science exposure (on each camera).
A single one-dimensional sky model is created by a b-spline fit to the airmass 
corrected sky spectra (See appendix b-spline for a continued description).
The airmass associated with the exposure mid-point (in time) is divided out 
of each sky spectrum (and errors scaled accordingly).  The differential airmass
across a single SDSS plate is the largest source of residuals after the dominant
wavelength residuals.  Individual pixels are masked if they were divided by a 
low flat field vector or a bad pixel was detected nearby (clarify?).
The number of discarded pixels before sky modeling is logged.

The sky

### 2005-02-03 doc/paper/skysubtraction.tex
latest update

@@
 
-A global sky model is fit using all available fiber spectra assigned as 
-"SKY" by target selection.  These fibers are set on blank fields, and are 
-used to estimate the sky level as a function of wavelength and CCD position 
-in individual exposures.   Each plate is nominally assigned 32 of its 640 
-fibers as "SKY", with the focal plane positions chosen to allow half of the 
-sky fibers to reach each spectrograph: 1 fiber per bundle of 20 is assigned to 
-"SKY".
+Sky subtraction is performed on a frame by frame basis using the 
+"flattened" wavelength calibrated extractions.  The goal of sky subtraction
+is to estimate the expected background counts contributing to each of 
+the native 320x2048 pixels in each extracted frame.  A "supersky" model 
+at a fiducial airmass of 1 is constructed by performing iterative fits to
+the extracted spectra associated with blank sky positions (so-called
+"sky" fibers).   The following steps are taken in 3 sequential iterations
+in order to produce the sky-subtracted spectra:
 
-The sky subtraction is done after the data have been extracted from 
-two-dimensional counts per CCD pixel to one-dimensional "corrected" electron 
-counts as a function of vacuum heliocentric wavelengths and fiber number.
-The flat-field "correction" removes differences due to surface brightness 
-changes due to varying spectral dispersion in different fibers.
+\begin{enumerate}
 
-A single subroutine is called three consecutive times to arrive at the final 
-model description for the sky background.  The first call uses all available sky
-fibers that were not masked as an entire fiber (e.g. by the fibermask bits).
-If only one good skyfiber is found, the routine logs an "ABORT" and returns.
-A median count level is summarized for the selected sky spectra, and the 5% 
-and 95% levels are reported for each science exposure (on each camera).
-A single one-dimensional sky model is created by a b-spline fit to the airmass 
-corrected sky spectra (See appendix b-spline for a continued description).
-The airmass associated with the exposure mid-point (in time) is divided out 
-of each sky spectrum (and errors scaled accordingly).  The differential airmass
-across a single SDSS plate is the largest source of residuals after the dominant
-wavelength residuals.  Individual pixels are masked if they were divided by a 
-low flat field vector or a bad pixel was detected nearby (clarify?).
-The number of discarded pixels before sky modeling is logged.
+\item{Identify the "sky" fibers which have no bits flagged 
+in the fiber mask array.}
 
-The sky
+\item{Divide each sky spectrum by the airmass of the fiber position appropriate
+for the midpoint of the observation to produce respresentative sky spectra for 
+an airmass of 1.  The inverse variance is scaled accordingly.}
+
+\item{Group and sort the sky spectra as a function of wavelength.  This step
+transforms a 2-dimensional set of N_sky x 2048 pixels into a sorted 1-dimensional array.  Each pixel maintains its associated central wavelength and 
+inverse variance. }
+
+\item{Perform a single cubic b-spline fit as a function of wavelength.  
+The number of breakpoints (number of free parameters is 4 greater) is set 
+at 2048 on the red side and 1536 on the blue side.  The breakpoint spacing 
+is set automatically to maintain approximately constant S/N between breakpoint 
+positions. The b-spline fit itself is iterative, with upper and lower rejection thresholds set to mask bad or deviant pixels. }
+
+\item{The b-spline fit is evaluated at all 320x2048 wavelengths representing all
+spectral pixels in the frame.  The airmass correction is applied for each individual fiber, and the rescaled super-sky is subtracted from each spectrum.}
+
+\item{Estimate the significance of the residuals in the sky fibers used for the
+fit in wavelength bins separated by breakpoints determined above.  for each bin
+where inverse the 67th percentile of $\chi^2$ is greater than 1, we rescale 
+all the inverse variance values down by the 67th percentile value of $\chi^$.
+The rescaling is actually done with a smooth interpolation as a function of 
+redshift with a constraint that the inverse variances cannot increase. }
+
+\end{enumerate}
+
+%The b-spline routines from the numerical fortran library: {\bf slatec}
+%were ported to IDL to be easily incorporated into the pipeline.
+
+The sequence followed in the determination of the sky subtracted spectrum
+and re-scaled errors is called 3 times successively.  After the 1st call,
+any sky fibers which have a median \chi^2 per pixel greater than 2 is 
+masked as a 'BADSKYFIBER' and the corresponding bit is set in the fibermask
+array.  If any fibers are masked, the sky subtraction is iterated a 2nd time.
+The sky subtraction routine is invoked for one final iteration, but with extra 
+freedom allowed in the fit.  The same freedom is allowed in wavelength, by
+keeping the same number of breakpoints, but a quadratic polynomial fit is
+allowed as a function of wavelength.  This triples the number of free 
+parameters in the linear least-squares fit, but the values are still 
+overconstrained due to the ample number of sky fibers.  The extra freedom
+in the model can accomodate variation in the PSF of sky features as a function 
+of position on the CCD.
+
+The sky subtraction method which utilizes the b-spline model takes advantage 
+of the inherent over-sampling in wavelength due to the large number of 
+sky fibers assigned per plate, without the requirement of resampling the 
+pixel value to a fixed linear wavelength scale.  The least squares fit
+is carried out for native spectral pixels where the errors are very nearly 
+uncorrelated and the fit is evaluated for all the native pixels (320x2048)  
+at the central wavelengths of the single spectral frames.
+
+The accuracy of the SDSS spectral sky subtraction has been tested in recent
+scientific analyses using large numbers of background-limited spectra.   
+The distribution of flux residuals between spectral observations 
+on separate nights has been shown to be 6\%-9\% larger in width than expected
+from the reported errors (McDonald et al. 2004, Burgess \& Burles 2005).
+For combined observations of faint galaxies, the distribution of residuals 
+about the best model fit is well described by a gaussian with exponential
+tails (Bolton et al. 2004), where the magnitude of tails is greatly enhanced in
+the vicinity of strong sky emission lines.

### 2005-02-03 doc/paper/paper2d.tex
Begin text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Definitions
%\pagestyle{empty}
\documentclass[12pt,preprint]{aastex}
%\documentstyle[emulateapj,danonecolfloat]{article}
%\usepackage{emulateapj,danonecolfloat}
\usepackage{rotating}

%------------------------------------------------------------------------------
% DEFINITIONS
%------------------------------------------------------------------------------
%
% Equations:
%
\def\BE{\begin{equation}}
\def\BEL#1{\begin{equation}\label{#1}}
\def\EE{\end{equation}}
%
% Random stuff:
%
\newcommand{\etal}{{\it et al.}~}
\newcommand{\LOGTEN}{{\log_{10}}}
%
% Units (math mode)
%
\newcommand{\Ang}{{\rm ~\AA}}
\newcommand{\cm}{{\rm ~cm}}
\newcommand{\dfdegree}{^\circ}
\newcommand{\degree}{^\circ}
\newcommand{\ergs}{{\rm ~erg~s}^{-1}}
\newcommand{\ergscmang}{{\rm ~erg~s}^{-1}{\rm cm}^{-2}{\rm\AA}^{-1}}
\newcommand{\ergcmHzsr}{{\rm ~erg~cm}^{-2}{\rm s}^{-1}
            {\rm Hz}^{-1}{\rm sr}^{-1}}
\newcommand{\Jy}{{\rm ~Jy}}
\newcommand{\JypSr}{{\rm ~Jy~sr^{-1}}}
\newcommand{\GHz}{{\rm ~GHz}}
\newcommand{\Hz}{{\rm ~Hz}}
\newcommand{\K}{{\rm ~K}}
\newcommand{\kms}{{\rm ~km/s}}
\newcommand{\MAG}{{\rm ~mag}}
\newcommand{\microK}{\mu{\rm K}}
\newcommand{\MJy}{{\rm ~MJy}}
\newcommand{\mJy}{{\rm ~mJy}}
\newcommand{\MJypSr}{{\rm ~MJy~sr^{-1}}}
\newcommand{\nWpMMSr}{{\rm ~nW~m}^{-2}{\rm sr}^{-1}}
\newcommand{\mK}{\rm ~mK}
\newcommand{\mm}{{\rm ~mm}}
\newcommand{\nMgy}{{\rm nMgy}}
\newcommand{\pc}{{\rm ~pc}}
\newcommand{\pix}{{\rm ~pix}}
\newcommand{\s}{{\rm ~s}}

%------------------------------------------------------------------------------
% TITLE PAGE
%------------------------------------------------------------------------------

\begin{document}

\title{The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline}

\author{
S. Burles\altaffilmark{\ref{MIT}}
and D. J. Schlegel\altaffilmark{\ref{Princeton}}
}
\altaffiltext{1}{Massachussetts Institute of Technology,
Boston, MA ????? \label{MIT}}
\altaffiltext{2}{Princeton University Observatory, Peyton Hall, Princeton,
NJ 08544 \label{Princeton}}

%------------------------------------------------------------------------------
% ABSTRACT
%------------------------------------------------------------------------------

\begin{abstract}
The Sloan Digital Sky Survey (SDSS) fiber-fed spectrographs have been used
to obtain spectra for $600,000$ galaxies, QSOs and stars.
We present the algorithms for reducing the raw data to
wavelength-calibrated, flux-calibrated one-dimensional spectra.
The wavelength calibration is shown to have an accuracy of $3 \kms$,
and the broad-band spectrophotometry has an accuracy of better than $10\%$.

%------------------------------------------------------------------------------
% SUBJECT HEADINGS
%------------------------------------------------------------------------------
\emph{Subject headings: }
???
\end{abstract}

%------------------------------------------------------------------------------
% INTRODUCTION
%------------------------------------------------------------------------------
\section{INTRODUCTION}
\label{sec_intro}
The Sloan Digital Sky Survey (SDSS) is a 5-band photometric survey of
approximately $10,000$ square degrees of the Northern sky and a concurrent
redshift survey of up to a million galaxies and 100,000 quasars
selected from the imaging survey \cite{york00}.  The primary 
science goals of the project are to provide the data to investigate the
large scale structure of the Universe and other extragalactic science.

The data reduction and products described herein are very similar to
those produced by the SDSS, with the most significant differences
in our treatment of spectro-photometric calibrations.
The data are released
with this paper and may be accessed via the WWW.\footnote{Full details
are at \texttt{http://spectro.princeton.edu}}
Future data releases will also be accessible at this site.

%------------------------------------------------------------------------------
%------------------------------------------------------------------------------
\section{OPTIMAL EXTRACTION}
\label{sec_extract}

%------------------------------------------------------------------------------
%------------------------------------------------------------------------------
\section{FLUX CALIBRATION}
\label{sec_fluxing}

The SDSS spectra are presented as spectro-photometric quantitites,
meaning that they have been converted to physical flux densities
as seen above the Earth's atmosphere.
There are many difficulties inherent in a fiber-fed spectrograph,
especially one with such a large field of view.
The problem is complicated by changes in the flat-field lamp spectra,
the throughput through some of the optical elements, changing
atmospheric transparency, and guiding errors.
Despite these problems, the spectrophotometry is accurate to
several percent on all wavelength scales.

Flux-calibration for the two spectrographs are treated almost completely
independently.  This is appropriate because the collimator optics
and CCDs are different.  Only the telescope optics are in common, and the
atmospheric extinction and guiding errors are continuous across a plate.
In general, the 320 fibers for spectrograph \#1 are plugged for those
objects south of the centerline of the field of view, and the 320 fibers
for spectrograph \#2 are plugged for the north.

The spectra are converted from raw counts (ADUs) to physical
units ($\ergs$).

$ 100 \ergscmang$

A mean relative response between fibers has already been corrected
using the quartz flat-field lamps (see Section ???).
Because of this flat-fielding, the remaining conversion from ADUs
to physical units should be identical for each spectrum on the 
same CCD, except for small differences from different pixel scales
and airmass.

The SDSS spectrophotometry is based upon our understanding
of F star subdwarfs.  These stars are the easiest main sequence
stars to theoretically model because ...???
For these reasons, we target 16 of these stars on each plate,
typically 8 on each spectrograph.

% See http://www.astro.princeton.edu:81/sdss-target/msg.725.html

We select relatively bright F dwarfs as spectrophotometric 
standards.  Ideally, we want the low-metallicity, halo stars in order 
to have spectra that have few spectroscopic features.
The F stars are chosen to satisfy the following color criteria:
$$ 0.6 < u-g < 1.2 $$
$$ 0.0 < g-r < 0.6 $$
$$ g-r > 0.75 * (u-g) - 0.45 $$
This selects stars that are on the blue edge of the main
sequence in (u-g) color.  At high Galactic latitudes,
??? stars per square degree fall in this color box.
Targetting priority is given to
those stars with colors closest to the following:
$$ u-g = 0.80 $$
$$ g-r = 0.30 $$
$$ r-i = 0.10 $$
This is meant to select those stars closest in color to BD +17 4708,
which is computed to have a similar color (0.93, 0.28, 0.01, courtesy
of David Hogg, priv. comm.).
From these stars, we select 8 on each plate in the magnitude
range $15.5 < g < 17.0$ (which we call SPECTROPHOTO_STD stars)
and 8 in the magnitude range $17.0 < g < 18.5$ (which we call
REDDEN_STD stars).
These magnitude limits are chosen to avoid nearing saturation
of the CCDs on the bright end, but to still ensure good S/N
on the faint end.
Assuming an absolute magnitude of 4.3 for these stars, this
implies distances of $1.7 < d < 3.5$ kpc for the SPECTROPHOTO_STD
stars and $3.5 < d < 6.9$ kpc for the REDDEN_STD stars.
Note that the color-selection and magnitude cuts are done on
``instrinsic'' magnitudes, meaning that we first apply corrections
for SFD extinction.

The stuff I'm calibrating to has been public for a while.  The original
ref is Gray & Corbally, 1994, AJ 107, 742.  There are some updates in
Gray, Graham, & Hoyt 2001, AJ, 121, 2159.  Gray's code SPECTRUM is
available http://www.acs.appstate.edu/dept/physics/spectrum/spectrum.html
You can't download the models -- you have to run them.  But I can send you
the ones I've run if you like.  (They start out as text, then I convolve
to SDSS resolution and make FITS files.)  Richard Gray was pretty nice
about resonding to email too.

%------------------------------------------------------------------------------
% CONCLUSIONS
%------------------------------------------------------------------------------
%\section{SUMMARY}

\newpage
%------------------------------------------------------------------------------
% REFERENCES
%------------------------------------------------------------------------------

\bibliographystyle{unsrt}
\bibliography{gsrp}

%\begin{thebibliography}{DUM}
%
%\bibitem[Schlegel, Finkbeiner \& Davis 1998 (SFD)]{sfd98} 
%Schlegel, D. J., Finkbeiner, D. P., \& Davis M. 1998, \apj, 500, 525 (SFD)
%
%\bibitem[York et al. 2000]{york00}
%York, D. G. \etal\ 2000, \aj, 120, 1579
%
%\end{thebibliography}
\clearpage

%------------------------------------------------------------------------------
%------------------------------------------------------------------------------
\appendix
\section{B-SPLINES}
\label{bsplines}

What the heck is a B-spline?

%------------------------------------------------------------------------------
%------------------------------------------------------------------------------
\appendix
\section{DATA MODEL}
\label{bsplines}

What the heck is a B-spline?

%------------------------------------------------------------------------------

\end{document}

### 2005-02-05 doc/paper/skysubtraction.tex
filled in some extraction text

@@
+\subsection{Extraction}
+
+We utilize an optimal extraction procedure to transform a 2-dimensional image of
+the fiber spectrograms to a set of 320 1-dimensional spectra.  We use the same procedure
+on all processed frames, and will describe the details in processing the flat-field frames, the
+arc lamp frames, and finally the science frames.  
+The extraction is performed row-by-row, so each frame requires an outside loop over 2048 individual rows.
+The counts in each row of 2048 pixels are modeled with a linear combination of 320 profiles plus a low-order
+polynomial background.  We decided to perform profile fitting to the SDSS spectral images 
+for the following reasons:
+
+\begin{enumerate}
+
+\item{The fiber profiles overlap significantly with their nearest neighbors.}
+
+\item{The profiles are relatively stable and smoothly vary over location on the CCDs.  
+They result from the convolution of the fiber illumination pattern + the point-spread function
+of the optics in the SDSS spectrographs + the window funciton of the Site CCD 24 micron square pixels. }
+
+\item{A least-squares fit with the true profile delivers an unbiased and minimum variance estimate of the counts
+(e.g. Horne et al. 1995).}
+
+\end{enumerate}
+
+The major obstacle to actual optimal extraction is the determination of the exact shape and position 
+of the underlying profile.  We investigated a range of symmetric, analytic profiles with a 
+minimum number of free parameters.  In the end, we found the best results using a simple gaussian 
+or a slightly steeper profile characterized by $e^{-|x|^3/\sigma^3}$.  The profiles
+are defined to be normalized to unity, so the amplitude of the profile fit gives a direct estimator
+of the counts at a single spectral position.
+
+Extraction is first carried out on all flat-field exposures (typically 10 second observations of
+4 quartz lamps projected on petals covering the entrance aperture of the telescope) which were observed
+with a common plug plate configuration.  The flat-field extraction is used to determine the shape of
+the profile as a function of wavelength and position on the CCD.  The resulting spectral extractions are also used to 
+determine the empirical fiber-to-fiber variations in throughput (as a function of wavelength) 
+during a single setup.
+
+Three types of exposures, corresponding to flat-field, arc lamp and science frames,
+ are extracted with the same basic procedure
+which we describe below.  The frames associated with a single observing sequence 
+are processed in a certain order.  The first types to be extracted are the 
+flat-fields frames ("FLATS").
+
 The object extraction lists 5 steps in header:
@@
 \item{Group and sort the sky spectra as a function of wavelength.  This step
-transforms a 2-dimensional set of N_sky x 2048 pixels into a sorted 1-dimensional array.  Each pixel maintains its associated central wavelength and 
-inverse variance. }
+transforms a 2-dimensional set of N_sky x 2048 pixels into a sorted 1-dimensional array.  
+Each pixel maintains its associated central wavelength and inverse variance. }
 
@@
 is set automatically to maintain approximately constant S/N between breakpoint 
-positions. The b-spline fit itself is iterative, with upper and lower rejection thresholds set to mask bad or deviant pixels. }
+positions. The b-spline fit itself is iterative, with upper and lower rejection 
+thresholds set to mask bad or deviant pixels. }
 
 \item{The b-spline fit is evaluated at all 320x2048 wavelengths representing all
-spectral pixels in the frame.  The airmass correction is applied for each individual fiber, and the rescaled super-sky is subtracted from each spectrum.}
+spectral pixels in the frame.  The airmass correction is applied for each individual 
+fiber, and the rescaled super-sky is subtracted from each spectrum.}
 
@@
 The distribution of flux residuals between spectral observations 
-on separate nights has been shown to be 6\%-9\% larger in width than expected
+on separate nights has been shown to be 8\% larger in width than the average expected
 from the reported errors (McDonald et al. 2004, Burgess \& Burles 2005).
-For combined observations of faint galaxies, the distribution of residuals 
-about the best model fit is well described by a gaussian with exponential
-tails (Bolton et al. 2004), where the magnitude of tails is greatly enhanced in
-the vicinity of strong sky emission lines.
+The spectra of faint, background-limited galaxies observed with SDSS exhibit a 
+distribution of residuals about the best fit spectral models described by the sum of a
+Gaussian and Laplace distribution (Bolton et al. 2004).  
+The influence of the exponential tail is greatly enhanced in the vicinity of strong sky 
+emission lines representative of systematic errors in the brightest lines, but the gaussian 
+width closely is closely represented by the reported variance array.

### 2005-02-05 doc/paper/paper2d.tex
Update.

@@
 
-The spectra are converted from raw counts (ADUs) to physical
-units ($\ergs$).
-
-$ 100 \ergscmang$
+% The spectra are converted from raw counts (ADUs) to physical
+% units ($\ergs$).
+% $ 100 \ergscmang$
 
@@
 standards.  Ideally, we want the low-metallicity, halo stars in order 
-to have spectra that have few spectroscopic features.
+to have spectra that have atmospheric opacities that are easy to model.
 The F stars are chosen to satisfy the following color criteria:
@@
 $$ g-r > 0.75 * (u-g) - 0.45 $$
+as shown in Figure ???.
 This selects stars that are on the blue edge of the main
@@
 ``instrinsic'' magnitudes, meaning that we first apply corrections
-for SFD extinction.
+for SFD extinction under the assumption that these stars are
+completely behind the Galactic dust.
+
+The spectra for these stars are compared to the model spectra
+from the SPECTRUM code of Gray \& Corbally (1994) as updated
+in Gray \etal\ (2001).
+We constructed a grid of 90 models with effective temperatures
+from $5000$ to $7000 \K$ (spaced every $250 \K$), [Fe/H] values
+from -2 to 0 (spaced every 0.5), and gravities of $\log g=4$
+and $\log g=4.5$.  These models are then ``observed'' as they
+would be seen in our spectra.  We convolve the models with
+the instrumental dispersion as a function of wavelength
+(see Section ???), and sample the results at the observed wavelengths.
+Since there are typically three exposures in the two spectrographs
+(blue and red) for each star, this means we re-generate the full
+grid of models for six observations of each star.
+We perform $\chi^2$ minimizations between these models and the
+6 spectra simultaneously.  We use the observed errors per pixel,
+which are independent because we have done no resampling of the data.
+All spectra and models are first median-filtered with a scale of $7000\kms$,
+and the regions of atmospheric absorption in the red are masked
+for the purpose of spectral typing our stars.
+A velocity in the range $-700 < v < +700 \kms$ is fit
+by comparing to the model with $T_{eff} = 6000\K$, Fe/H=$-1.5$,
+$\log g=4$.  All the models are redshifted with this best-fit velocity,
+and the resulting model with the smallest $\chi^2$ is chosen
+as the best-fit model.  We reject any stars with a reduced $\chi^2$
+larger than 2.0 in the full spectrum, with a reduced $\chi^2$
+larger than 2.0 in the wavelength regions of the strong stellar
+absorption lines, or with a median S/N less than 2.0 in
+those absorption line regions.
+We also reject stars if more than $20\%$ of the pixels are bad
+in any of the 3 or more observations, which typically only happens
+if a fiber is broken or the spectrum falls on bad columns on the CCD.
+Figure ??? shows the distribution of parameters that we have fit
+for the ??? flux-calibration stars on the Data Release 4 plates.
+
+Each flux-calibration star provides a means of converting our
+observed spectra to physical units.  
+In practice, a single star is not sufficient because of the
+slight undersampling in wavelength, bad pixels in any
+particular observation, stellar typing errors,
+and the desire to achieve better S/N in the flux-calibration
+vector than one star provides.
+
+Models are redshifted, fluxed, reddened.
 
@@
 
-%\begin{thebibliography}{DUM}
-%
-%\bibitem[Schlegel, Finkbeiner \& Davis 1998 (SFD)]{sfd98} 
-%Schlegel, D. J., Finkbeiner, D. P., \& Davis M. 1998, \apj, 500, 525 (SFD)
-%
-%\bibitem[York et al. 2000]{york00}
-%York, D. G. \etal\ 2000, \aj, 120, 1579
-%
-%\end{thebibliography}
+\begin{thebibliography}{DUM}
+
+%Gray's code SPECTRUM is
+%available http://www.acs.appstate.edu/dept/physics/spectrum/spectrum.html
+\bibitem[Gray et al. 2001]{gray01}
+Gray, R.\ O., Graham, P.\ W., \& Hoyt, S.\ R. 2001, \aj, 121, 2159
+
+\bibitem[Gray \& Corbally 1994]{gray94}
+Gray, R.\ O., \& Corbally, C.\ J. 1994, \aj, 107, 742
+
+\bibitem[Schlegel, Finkbeiner \& Davis 1998 (SFD)]{sfd98} 
+Schlegel, D.\ J., Finkbeiner, D.\ P., \& Davis M.\ 1998, \apj, 500, 525 (SFD)
+
+\bibitem[York et al. 2000]{york00}
+York, D.\ G. \etal\ 2000, \aj, 120, 1579
+
+\end{thebibliography}
 \clearpage

### 2005-03-09 doc/paper/paper2d.tex
Make TeXable.

@@
 From these stars, we select 8 on each plate in the magnitude
-range $15.5 < g < 17.0$ (which we call SPECTROPHOTO_STD stars)
+range $15.5 < g < 17.0$ (which we call SPECTROPHOTO\_STD stars)
 and 8 in the magnitude range $17.0 < g < 18.5$ (which we call
-REDDEN_STD stars).
+REDDEN\_STD stars).
 These magnitude limits are chosen to avoid nearing saturation
@@
 Assuming an absolute magnitude of 4.3 for these stars, this
-implies distances of $1.7 < d < 3.5$ kpc for the SPECTROPHOTO_STD
-stars and $3.5 < d < 6.9$ kpc for the REDDEN_STD stars.
+implies distances of $1.7 < d < 3.5$ kpc for the SPECTROPHOTO\_STD
+stars and $3.5 < d < 6.9$ kpc for the REDDEN\_STD stars.
 Note that the color-selection and magnitude cuts are done on
@@
 The stuff I'm calibrating to has been public for a while.  The original
-ref is Gray & Corbally, 1994, AJ 107, 742.  There are some updates in
-Gray, Graham, & Hoyt 2001, AJ, 121, 2159.  Gray's code SPECTRUM is
+ref is Gray \& Corbally, 1994, AJ 107, 742.  There are some updates in
+Gray, Graham, \& Hoyt 2001, AJ, 121, 2159.  Gray's code SPECTRUM is
 available http://www.acs.appstate.edu/dept/physics/spectrum/spectrum.html

### 2006-03-22 bin/spredrobot.sh
spsyncrobot.sh always lists all nights it scans. Make sure we do not process nights _too_ many times.

@@ -46,7 +46,7 @@ echo "IDLUTILS_DIR="$IDLUTILS_DIR
 #------------------------------------------------------------------------------
 # Find copied mjds which have not been reduced
 
-mjds=`fgrep -v -f $plannedMJDs $copiedMJDs`
+mjds=`fgrep -v -f $plannedMJDs $copiedMJDs | sort -n | uniq`
 mjdlist=`echo $mjds | perl -ane 'print join(",",split())'`
 
 #------------------------------------------------------------------------------

### 2006-03-22 bin/spsyncrobot.sh

@@ -91,7 +91,7 @@ mjdlist=''
 # The following line fails, because the directory list too long!
 # So instead, "cd" into the directory first.
 # remotedir=`$SPROBOT_RSH $hostname ls -d /data/spectro/astrolog/[5-9]???? | tail -7 | sed 's/\/data\/spectro\/astrolog\///g' | sed 's/\///g'`
-remotedir=`$SPROBOT_RSH $hostname "(cd /data/spectro/astrolog ; ls -d [5-9]???? | tail -20 | sed 's/\///g')"`
+remotedir=`$SPROBOT_RSH $hostname "(cd /data/spectro/astrolog ; ls -d [5-9]???? | tail -7 | sed 's/\///g')"`
 echo REMOTEDIR=$remotedir
 
 for mjdstr in $remotedir ; do

### 2006-03-24 doc/paper/skysubtraction.tex
overdue for checkin

@@
 during a single setup.
+
+The first very critical step is to robustly determine the fiber positions in CCD pixel coordinates.
+This is performed by identifying peaks in a cross-sectional cut of across the center row (actually a median
+cross-section of the central 9 rows).  The correct identification of the 320
+fibers per frame can be uniquely determined by demanding a wider spacing between the bundles (about 10 pixels)
+relative to the nominal inter-fiber spacing (6.1 pixels on average).  Once the starting peaks and 
+identification have been determined, we trace the fiber centers by performing flux-weighted centroids on subsequent rows
+using the previous row's center as the intial guess.  This has proven to work spectacularly well on SDSS
+flat-fields, and we have yet to uncover a failure mode in over 10000 flat-field frames.
+The flux-weighted centroids do have scatter from shot noise, so we fit a 7th order polynomial to each set of
+fiber centroids, and the best-fit coefficients are stored to represent the fiber positions on each
+flat-field frame.
 

### 2006-03-28 doc/paper/skysubtraction.tex
concurrent update

@@
 We utilize an optimal extraction procedure to transform a 2-dimensional image of
-the fiber spectrograms to a set of 320 1-dimensional spectra.  We use the same procedure
-on all processed frames, and will describe the details in processing the flat-field frames, the
+the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
+We use the same procedure on all processed frames, 
+and will describe the details in processing the flat-field frames, the
 arc lamp frames, and finally the science frames.  
-The extraction is performed row-by-row, so each frame requires an outside loop over 2048 individual rows.
-The counts in each row of 2048 pixels are modeled with a linear combination of 320 profiles plus a low-order
-polynomial background.  We decided to perform profile fitting to the SDSS spectral images 
-for the following reasons:
+The extraction is performed row-by-row, so each frame requires an 
+outside loop over 2048 individual rows.
+The counts in each row of 2048 columns are modeled with a linear 
+combination of 320 profiles plus a low-order
+polynomial background.  In addition, deviations to the trace centers and
+the profile widths are calculated as linear basis modes (reflecting the first
+and second spatial derivatives, respectively).  The compiled routine
+we implement to fit the profiles in a single row (extract_row.c) can
+be invoked with any combination of these parameters.
+
+We decided to perform profile fitting to the 
+SDSS spectral images for the following reasons:
 
@@
 
-\item{The profiles are relatively stable and smoothly vary over location on the CCDs.  
-They result from the convolution of the fiber illumination pattern + the point-spread function
-of the optics in the SDSS spectrographs + the window funciton of the Site CCD 24 micron square pixels. }
+\item{The profiles are relatively stable and smoothly vary over 
+location on the CCDs.  They result from the convolution of the fiber 
+illumination pattern + the point-spread function
+of the optics in the SDSS spectrographs + the window funciton of the 
+Site CCD 24 micron square pixels. }
 
-\item{A least-squares fit with the true profile delivers an unbiased and minimum variance estimate of the counts
-(e.g. Horne et al. 1995).}
+\item{A least-squares fit with the true profile delivers an 
+unbiased and minimum variance estimate of the counts (e.g. Horne et al. 1995).}
 
@@
 
-The major obstacle to actual optimal extraction is the determination of the exact shape and position 
-of the underlying profile.  We investigated a range of symmetric, analytic profiles with a 
-minimum number of free parameters.  In the end, we found the best results using a simple gaussian 
-or a slightly steeper profile characterized by $e^{-|x|^3/\sigma^3}$.  The profiles
-are defined to be normalized to unity, so the amplitude of the profile fit gives a direct estimator
-of the counts at a single spectral position.
+The major obstacle to actual optimal extraction is the determination of 
+the exact shape and position of the underlying profile.  
+We investigated a range of symmetric, analytic profiles with a 
+minimum number of free parameters.  In the end, we found the best 
+results using a simple Gaussian or a slightly steeper profile 
+characterized by $e^{-|x|^3/\sigma^3}$.  The profiles are defined to be 
+normalized to unity, so the amplitude of the profile fit is a 
+direct estimator of the counts at a single spectral position.
 
-Extraction is first carried out on all flat-field exposures (typically 10 second observations of
-4 quartz lamps projected on petals covering the entrance aperture of the telescope) which were observed
-with a common plug plate configuration.  The flat-field extraction is used to determine the shape of
-the profile as a function of wavelength and position on the CCD.  The resulting spectral extractions are also used to 
-determine the empirical fiber-to-fiber variations in throughput (as a function of wavelength) 
-during a single setup.
+Extraction is first carried out on all flat-field exposures 
+(typically 10 second observations of 4 quartz lamps projected on petals covering the entrance aperture of the telescope) which were observed
+with a common plug plate configuration.  The flat-field extraction is 
+used to determine the shape of
+the profile as a function of wavelength and position on the CCD.  
+The resulting spectral extractions are also used to 
+determine the empirical fiber-to-fiber variations in throughput 
+(as a function of wavelength) during a single setup.   
+A flat-field frame is labeled rejected by the pipeline if one of the
+following conditions is met:  1) More the 2\% of the CCD pixels are labeled
+bad by sdssproc, 2) More than 100 rows of the image show saturation, 
+3) The 80th percentile count level is less than 1000 counts, 
+4) Fewer than all 8 flat-field telescope petals were closed, 
+5) Fewer than all 4 flat-field lamps were on,
+6) More than 10000 pixels were rejected during fiber tracing, 
+7) Fiber tracing fails, or traces are seperated by less than 3 pixels anywhere
+on the flat-field image.
 
-The first very critical step is to robustly determine the fiber positions in CCD pixel coordinates.
-This is performed by identifying peaks in a cross-sectional cut of across the center row (actually a median
-cross-section of the central 9 rows).  The correct identification of the 320
-fibers per frame can be uniquely determined by demanding a wider spacing between the bundles (about 10 pixels)
-relative to the nominal inter-fiber spacing (6.1 pixels on average).  Once the starting peaks and 
-identification have been determined, we trace the fiber centers by performing flux-weighted centroids on subsequent rows
-using the previous row's center as the intial guess.  This has proven to work spectacularly well on SDSS
-flat-fields, and we have yet to uncover a failure mode in over 10000 flat-field frames.
-The flux-weighted centroids do have scatter from shot noise, so we fit a 7th order polynomial to each set of
-fiber centroids, and the best-fit coefficients are stored to represent the fiber positions on each
-flat-field frame.
+The first very critical step is to robustly determine the fiber positions 
+in CCD pixel coordinates.
+This is performed by identifying peaks in a cross-sectional cut of across 
+the center row (actually a median cross-section of the central 9 rows).  
+The correct identification of the 320 fibers per frame can be 
+uniquely determined by demanding a wider spacing between the bundles 
+of 20 fibers (spaced by about 10 pixels) relative to the nominal inter-fiber 
+spacing (6.1 pixels on average).  Once the starting peaks and 
+identification have been determined, we trace the fiber centers 
+by performing flux-weighted centroids on subsequent rows
+using the previous row's center as the intial guess.  This has proven to 
+work well on SDSS flat-fields, and we have yet to uncover a failure 
+mode in over 10000 flat-field frames.
+The flux-weighted centroids do have scatter from shot noise, 
+so we fit a 7th order polynomial to each set of fiber centroids, 
+and the best-fit coefficients are stored to represent the fiber positions 
+on each flat-field frame.
 
-Three types of exposures, corresponding to flat-field, arc lamp and science frames,
- are extracted with the same basic procedure
-which we describe below.  The frames associated with a single observing sequence 
-are processed in a certain order.  The first types to be extracted are the 
-flat-fields frames ("FLATS").
+Each flat-field frame is extracted twice, once assuming a Gaussian
+profile and again assuming the steeper $|x|^3$ profile.  In both cases,
+the width is allowed to vary freely.  The total number of free parameters
+per row in the flat-field extractions is equal to 2 per fiber plus 10
+parameters to represent a smooth polynomial background.  After the extractions
+are complete, $\chi^2$ is computed by directly comparing the model fit
+to the flat-field image.  The median $\chi^2$ for each case is compared
+directly, and the better fit and profile (lower median $\chi^2$) are retained
+and implemented for the extraction of the remaining frames.  Also retained
+is a smooth polynomial fit the average profile width in each fiber bundle
+as a function of CCD row number.
 
-The object extraction lists 5 steps in header:
+After all flat-field frames for a given sequence have been extracted
+and stored, the arc-lamp frames in the same sequence are extracted.  
+Each arc-lamp frame is matched with the nearest (in time) 
+non-rejected flat-field frame.  The matched flat-field gives the 
+initial trace centers to
+be assumed, as well as the profile shape and width determined from the
+flat-field.  To account for any flexure between the frames, the trace
+centers are allowed to shift/stretch to the arc-lamp image given a 
+3x3 polynomial distortion kernel over the entire extent of the image
+(match_trace.pro).
+
+The arc-lamp frames are extracted row-by-row, in the same
+manner as the flat-fields, the number of free parameters assumed
+for the model fit is 2 per fiber (profile area and width) and 1 parameter
+to fit for a constant background per row.  The restricted freedom in the
+background parameterization reflects the lower scattered light levels in
+the images of the arc-lamps.
+
+\subsection{(or arc-lamp spectra -> wavelength map)}
+
+We determine an initial wavelength solution for every arc-lamp frame
+successfully extracted in the following manner:
+
+For the appropriate side of each spectrograph (red or blue), we 
+construct a list of suitable unblended lines from a parent list including
+known lines of Mercury, Cadmium, Neon and (a trace of) Argon 
+that are readily observable in the standard set of SDSS arc-lamps. 
+A representative spectrum is calculated as the median extracted spectrum
+of the 5 closest "unflagged" spectra falling near the central columns 
+of the arc-lamp image.  This median arc-spectrum is matched with 
+cross-correlation to a simulated spectrum of suitable arc-line positions and 
+strengths by systematically searching a small volume of 
+parameter space which conservatively covers
+the polynomial coefficients of the SDSS wavelength solution.
+In the first iteration, we search a set of 
+100 coefficient vectors allowing for quadratic changes and return 
+the best match.  In the second iteration, we search in a smaller region
+about the previous best match, and search a list of 125 coefficient vector
+allowing for cubic deviations.  The coefficient vector with the best
+match from the second iteration is returned as an initial guess along with
+the cross-correlation coefficient.
+
+A 2-step iterative procedure is loop through
+a range of possible polynomial
 

### 2006-03-28 doc/paper/paper2d.tex
Update.

@@
 
+We first convert from ADUs to physical flux units under the
+assumption that there are no guiding or object centroiding errors.
 A mean relative response between fibers has already been corrected
@@
 to physical units should be identical for each spectrum on the 
-same CCD, except for small differences from different pixel scales
+same CCD, except for small, known differences from different pixel scales
 and airmass.
@@
 Figure ??? shows the distribution of parameters that we have fit
-for the ??? flux-calibration stars on the Data Release 4 plates.
+for the ??? flux-calibration stars through Data Release 5.
 
@@
 slight undersampling in wavelength, bad pixels in any
-particular observation, stellar typing errors,
+particular observation, uncertainties in our stellar typing,
 and the desire to achieve better S/N in the flux-calibration
-vector than one star provides.
+vector than one star provides.  Each of these F star spectra
+are ratioed with the best-fit model at the best-fit redshift,
+reddened with the SFD reddening maps, and scaled to the SDSS r-band
+magnitudes.  The reddening is applied using an $R_V=3.1$ extinction
+law from O'Donnell (1994), and assuming that these stars reside
+behind all the Galactic dust.
+These ratios are called the ``flux calibration vectors.''
 
-Models are redshifted, fluxed, reddened.
+The set of flux-calibration vectors for an entire plate are combined
+using B-splines to compute an average flux-calibration vector per CCD.
+For the red CCDs, we fit 2-dimensional B-splines where the fits are
+allowed to vary linearly with airmass at each spline break point
+in wavelength.  These airmass terms are disabled if the plate spans
+less than 0.1 in airmass, or if there are fewer than 3 good stars.
+Because the B-spline fits can be dominated by
+spectrophotometry errors of low order in wavelength, and by
+atmospheric throughput variations, each fluxing
+vector is first corrected to the mean vector using cubic polynomials.
+The B-spline fits are then evaluated on an object-by-object basis,
+using the airmass of each fiber, and multiplying back in the mean
+cubic correction for the F stars each exposure.
+The mean of those cubic corrections are computed such that they
+are consistent for the same objects between the blue and red CCDs
+on each exposure.
 
-The stuff I'm calibrating to has been public for a while.  The original
-ref is Gray \& Corbally, 1994, AJ 107, 742.  There are some updates in
-Gray, Graham, \& Hoyt 2001, AJ, 121, 2159.  Gray's code SPECTRUM is
-available http://www.acs.appstate.edu/dept/physics/spectrum/spectrum.html
-You can't download the models -- you have to run them.  But I can send you
-the ones I've run if you like.  (They start out as text, then I convolve
-to SDSS resolution and make FITS files.)  Richard Gray was pretty nice
-about resonding to email too.
+At this point, the spectra can be flux-calibrated to the mean
+F star response in each exposure, including correction for
+atmospheric absorption on an object-by-object basis.
+The problem would be solved at this point if there were
+no centroiding errors, guiding errors, differential refraction,
+or transparency variations across the 3-degree field.
+...???
+
+%------------------------------------------------------------------------------
+\subsection{GALACTIC REDDENING}
+
+The SDSS spectra are presented as observed above the Earth's
+atmosphere, but not above the Galaxy's atmosphere.
+If one is studying distant stars or extragalactic objects,
+one may wish to remove the effects of Galactic extinction
+on these spectra.  It is suggested that one use the SFD reddening
+maps and the extinction curves from either Cardelli \etal\ (1989)
+or O'Donnell (1994) with $R_V=3.1$.  The differences between the
+two extinction curves is less than $1\%$, but we choose the O'Donnell
+curves as the more modern values.  For the optical wavelengths
+of interest, the relevant formulae are
+assembled from those two papers below:
+$$ {A_\lambda \over A_V} = a(y) + {b(y)\over R_V} $$
+$$ y \equiv {10^4\AA \over \lambda} - 1.82 $$
+$$ a(y) = 1.000 + 0.104 y - 0.609 y^2 + 0.701 y^3 + 1.137 y^4
+  - 1.718 y^5 - 0.827 y^6 + 1.647 y^7 - 0.505 y^8 $$
+$$ b(y) = 1.952 y + 2.908 y^2 - 3.989 y^3 - 7.985 y^4
+  + 11.102 y^5 + 5.491 y^6 - 10.805 y^7 + 3.347 y^8 $$
+
+{\bf Need CCM above 9091 Ang!!!???}
+
+Note that the extinction corrections should be applied in the
+observed wavelength frame, before any de-redshifting of extragalactic objects.
 
@@
 Gray, R.\ O., \& Corbally, C.\ J. 1994, \aj, 107, 742
+
+\bibitem[O'Donnell 1994]{odonnell94} 
+O'Donnell, J.\ E. 1994, \apj, 422, 158
 

### 2006-03-28 doc/paper/skysubtraction.tex
another small update

@@
 
-A 2-step iterative procedure is loop through
-a range of possible polynomial
+We apply the best guess coefficients to predict the approximate pixel 
+positions of the suitable list of arc lines.  The arc line centroids
+are traced across extracted spectra (along the axis of 320 fibers).
+The centroids are fit to a smooth polynomial as a function of fiber
+position on the arc-lamp image, and subsequently fit through 6 iterations
+with simple flux weighting to arrive at a converged measure of line 
+centroids in each extracted spectrum.  Lines are excluded from the resulting
+list if the flux-weighted centroids are bad in more than 10\% of the
+spectra or if the RMS residuals are larger than 20\% of a pixel.
+We perform an iterative 4th order Legendre polynomial fit of 
+logarithmic wavelength as a function of pixel in each spectrum, 
+and retain a list of rejected centroids, 
+which have deviations relative to the model of
+10\% of a pixel width.  An additional requirement is applied to each
+suitable arc line, wherein the line is rejected in no more than 3 fibers
+per bundle of 20.  Rejected centroids are replaced with the appropriate
+positions based on a linear interpolation of the differences of non-masked
+centroids relative to updated polynomial fits per spectrum.
+The final set of positions, where the same number of line positions have
+reliable centroids per spectrum, are fit to produce the final wavelength
+coefficients which give the logarithmic (base 10) wavelengths as a function
+of spectral pixel.
+
+An arc-lamp image is declared bad using the first 4 tests as employed
+in tests of the flat-field images.  In addition, an arc-lamp exposure
+is declared unacceptable ("BAD") if any of these following criteria are met:
+5) All 8 arc-lamps must be on, 4 Neon lamps and 5 Mercury-Cadmium lamps,
+6) The initial cross-correlation coefficient is less than 0.5,
+7) Less than 6 arc lines are acceptable after quality assurance checks.
+
+The width of unblended arc lines are determined as an average over
+each bundle of 20 spectra.  In short, the arc-lamp emission lines 
+are extracted (with a Gaussian spectral profile) 
+allowing free parameters for the area and width of each suitable line.
+The resulting Gaussian widths per fiber bundle are fit 
+with a low order polynomial as a function of average line position
+(the fit is quadratic and cubic for the blue and red side respectively).
+The fitted quantity is misnomerly (sp?) referred to as the arc-lamp dispersion,
+although the values reflect the standard deviation of a Gaussian spectral
+profile in pixel units.
+
+(or just refer to data model appendix instead of the paragraph below:)
+
+The arc-lamp reductions are written to disk at this stage, and
+stored in an extended FITS file:  HDU #0 contains the processed
+and extracted arc-lamp spectra as counts per pixel, HDU #1 contains
+the identified arc line wavelengths and the measured line centroids
+(including replaced positions of rejections), HDU #2 contains a single
+structure with the Legendre polynomial coefficients applicable to each
+of the 320 arc-lamp spectra, HDU #3 contains the maskbits for each spectrum,
+and HDU #4 contains the polynomial coefficients to describe the arc-lamp
+Gaussian widths as a function of spectral pixel.
 

### 2006-03-29 doc/paper/skysubtraction.tex
Tuesday's update

@@
 
+\item{A model profile fit leads to a model image of the SDSS data,
+and iteration allows cosmic-ray rejection and goodness-of-fit tests.}
+
 \end{enumerate}
@@
 (typically 10 second observations of 4 quartz lamps projected on petals covering the entrance aperture of the telescope) which were observed
-with a common plug plate configuration.  The flat-field extraction is 
-used to determine the shape of
+with a common plug plate configuration (a spectroscopic sequence). 
+The flat-field extraction is used to determine the shape of
 the profile as a function of wavelength and position on the CCD.  
@@
 determine the empirical fiber-to-fiber variations in throughput 
-(as a function of wavelength) during a single setup.   
-A flat-field frame is labeled rejected by the pipeline if one of the
+as a function of wavelength.   
+A flat-field frame is rejected by the pipeline if one of the
 following conditions is met:  1) More the 2\% of the CCD pixels are labeled
@@
 The final set of positions, where the same number of line positions have
-reliable centroids per spectrum, are fit to produce the final wavelength
+reliable centroids per spectrum, are fit with equal weighting per line
+position to produce the final wavelength
 coefficients which give the logarithmic (base 10) wavelengths as a function
@@
 
-An arc-lamp image is declared bad using the first 4 tests as employed
+An arc-lamp image is declared unacceptable using the first 4 tests as employed
 in tests of the flat-field images.  In addition, an arc-lamp exposure
@@
 (the fit is quadratic and cubic for the blue and red side respectively).
-The fitted quantity is misnomerly (sp?) referred to as the arc-lamp dispersion,
-although the values reflect the standard deviation of a Gaussian spectral
-profile in pixel units.
+The fitted quantity is referred to as the arc-lamp dispersion,
+but the values reflect the standard deviation of a Gaussian 
+profile in pixel units and represents the line-spread function of the 
+arc-lamp spectra.
 
@@
 Gaussian widths as a function of spectral pixel.
+
+\subsection{Constructing fiber flats}
+
+The final step in the reduction of the calibration frames in a given
+sequence is the construction of the fiber flat vectors.  Each flat-field
+frame is matched to the nearest arc-lamp frame as before, and the
+final flat-field fiber spectra are extracted in two steps.  The first
+step uses the preferred spatial profile and width determined above,
+and the extraction procedure produces a model of the flat-field frame.
+The model is used to subtract a diffuse scattered background given
+by an exponential point-spread function of :  (CCD halo PSF).
+We apply a 1-dimensional version of this 2-d PSF as an approximation,
+and allow the PSF to vary appropriately with wavelength.
+The final extraction is applied to the scattered light subtracted flat-field
+image allowing for 2 free paramters per fiber per row and 
+5 background parameters per row.
+
+The final flat-field spectra are stored as 
+detected counts per fiber per pixel with wavelengths of each pixel
+given by the associated solution of the arc-lamp frame.  
+We model the flat-field spectra as the product:
+
+\begin{equation} 
+{\rm Flat Model}_{ij} = A * f_{avg}(\lambda) \, {\rm fflat_i} (\lambda) 
+\end{equation}
+
+where $A$ is a overall normalization in counts, 
+$f_{avg}$ is the average response per constant logarithmic wavelength
+interval, $fflat_i$ is the fiber-to-fiber response as a function of wavelength.
+This model of the extracted flat-field spectra provides a transformation to
+compare detected counts from one fiber to another.
+
+This should remove the differential response due to the individual fiber
+throughput, as well as the small relative differences in the size of the
+spectral pixels at a constant wavelength.  ??Why do we correct for pixel
+sizes later on if the differential wavelength dispersion is included in fflat??
+
+The common normalized response function, $f_{avg}$, is specified as a B-spline
+function with breakpoints spaced approximately every native pixels.
+An example of normalized individual fiber spectra 
+and the resulting B-spline fit are shown in Figure ?.  The individual fiber
+flat vectors are fits to the ratio of the extracted spectra to the superflat
+(define above).  The individual vector fits are also B-splines with breakpoints
+spaced approximately every 5 native pixels.  The final set of flat vectors are
+stored as an image with dimensions Nfiber x Npixels.
+
+\subsection{Object Extraction}
+
+With the extracted calibration frames for a given spectroscopic sequence
+reduced, we proceed to extract the remaining science frames.
+%To minimize the differences between individual science frames
+We first select a set of arc-lamp and flat-field frames as the common
+choice of initial calibrations for all science frame extractions in a given
+observing sequence.  We loop through the science exposures, processing the
+raw images with bias and pixel-to-pixel flat images (see section ??).
+We reject a science exposure outright if 1) more than 10\% of the pixels
+are flagged as bad, 2) more than 100 rows are saturated, 3) the 25th
+percentile count level is greater than 1000 counts (the nominal level
+is of order 10 counts), 4) any of the flat-field petals are closed, or 5)
+any of the calibration lamps are recorded as on.
+
+Inputs to object extraction include the wavelength coefficients derived
+from the associated arc-lamp exposure, both the superflat and fiber flat
+vectors from the selected flat-field exposure, the profile shape and width
+as a function of position on the CCD image.
+
+The first step in object extraction is a simple tophat aperture (boxcar) 
+extraction centered at the fiber positions determined from the flat-field 
+and with a full spatial width of 6 CCD pixels.  The median counts per fiber 
+are passed directly to a test for overly bright fibers. Fiber spectra are 
+identified as bright (a.k.a. "WHOPPING") if the median counts per pixel in the 
+boxcar extraction is greater than 10000 counts above the median counts in the 
+frame.  In addition, only the brightest fiber in a spatially continuous set of 
+bright fibers is labelled "WHOPPING".
 

### 2006-03-29 doc/paper/skysubtraction.tex
Wednesday's lunch update

@@
 identified as bright (a.k.a. "WHOPPING") if the median counts per pixel in the 
-boxcar extraction is greater than 10000 counts above the median counts in the 
-frame.  In addition, only the brightest fiber in a spatially continuous set of 
+boxcar extraction are greater than 10000 counts above the median in the 
+frame.  In addition, only the brightest fiber in a spatially contiguous set of 
 bright fibers is labelled "WHOPPING".
 
+A pixelmask vector is created for each fiber spectrum to be extracted.
+See Appendix ?? for a description of the mask bits and their use.
+
+We fit for a slight distortion of the stored flat-field fiber traces to 
+best match the fiber positions of the science image.  Nominal shifts and
+distortions are usually less than a native CCD pixel.  The science frame
+is now extracted in a series of 6 steps which we detail below.
+
 \begin{enumerate}
-\item{Locate bright fibers (aka whopping fibers)}
-\item{Assess sky fiber levels}
-\item{Commence iterative Optimal extraction }
-\item{Shift wavelength solution to match sky lines (describe in wavelength solution)}
-\item{Sky subtract}
-
+\item{Initial extraction including broad wings of bright (WHOPPING) fibers.}
+\item{Reconstruct a smooth background light model based on fitted parameters.}
+\item{Calculate a scattered halo image using the model of the fiber spectra less
+the aforementioned smooth background.}
+\item{Re-extract fiber spectra on the scattered halo subtracted image.}
+\item{Reconstruct smooth background and subtract.}
+\item{Final extraction with no free parameters in the background and
+no free parameters in the profile widths.}
 \end{enumerate}
 
-Note: Flux correction and Telluric correction have been moved (a long time ago).
+The extracted science spectra are first normalized by the fiber flat vectors.
+Strong atmospheric emission lines (sky lines) are identified in the the 
+set of spectra, and the wavelength solution is shifted to match the 
+identified lines.  Examples of the shift corrections are shown in Figure ??.  
+The widths of the identified sky lines are also measured with the same method
+as the arc-line width measurements.  All spectra are also normalized by
+the average flat-field spectrum (the superflat) to remove small scale
+features likely attributed to the chromatic response of the dichroic.
+The wavelength solution is corrected to both vacuum wavelengths and the
+heliocentric velocity frame.
 

### 2006-03-29 doc/paper/paper2d.tex
Update.

@@
 
+\subsection{Flux calibration to F stars}
+
 The SDSS spectrophotometry is based upon our understanding
@@
 or transparency variations across the 3-degree field.
-...???
+We assume that these remaining fluxing problems are all
+low-order functions with wavelength.
+
+\subsection{Flux correction vectors}
+
+The co-adding of spectra from multiple exposures demands
+consistency of the spectrophotometry at high precision between those
+exposures, more so than the overall accuracy.
+For this reason, we choose a ``best'' exposure for each plate
+to which the other exposures are corrected on an object-by-object basis.
+The best exposure is chosen to be that whose worst CCD is higher S/N
+than the worst CCD in any other exposure.
+We loop over each fiber on each camera, and solve for the multiplicative,
+cubic function that remaps each exposure to the best exposure.
+This need be done carefully to avoid any remaining bad pixels,
+such as cosmic rays, from throwing the fit.
+The fit is first performed by fixing the errors per pixel, and only
+rescaling the spectrum from the exposure being compared to the best exposure.
+This is a linear problem which we iterate, rejecting 2.5-sigma outliers.
+(These points will not necessarily be rejected from the final spectra.)
+The flux-correction vectors are then re-fit as a non-linear problem,
+where the per-pixel errors are rescaled with the spectra.
+Some of the spectra, such as for sky fibers, do not warrant high-order
+polynomial corrections.  For this reason, we perform fits to the
+flux-correction vectors with increasing polynomial orders.
+If the next-higher order does not improve the $\chi^2$ of the fit
+by at least 5, then we stop the fit at that order.  The highest-order
+fits allowed are cubic.
+
+\subsection{Co-adding the spectra with B-splines}
+
+The set of all blue and red exposures for each fiber are
+combined using a B-spline.
+The splines are evaluated on a wavelength grid that
+is uniform is log-wavelengths, with the scale always chosen
+as $10^{-4}$ in log-wavelength.  The wavelength ranges are
+slightly different for each plate, chosen to span the input data.
+
+\subsection{Flux distortion image}
+
+The remaining spectro-photometric corrections are functions across
+the entire plate that are designed to remove large-spatial scale centering
+and guiding errors.
+Some of the terms are achromatic with position on the focal plane
+(X=XFOCAL and Y=YFOCAL).  There are also chromatic terms that scale
+as $ L \equiv 1 - (5100\AA / \lambda)^2 $, since that function gives
+an equal effect between $3900$ and $5070\AA$ as between $5070$
+and $9000\AA$.
+There are also magnitude offsets as a function fo spectrograph ID,
+and a chromatic offset as a function of spectrograph ID.
+$$ F' = F  \times \left[ a_0 ({\rm SPECID~EQ~1}) + a_1 ({\rm SPECID~EQ~2}) \right]
+  \times \exp \left[ a_2 x + a_3 y + a_4 xy + a_5 x^2 + a_6 y^2 + a_7 xL + a_8 yL
+  + a_9 L ({\rm SPECID~EQ~1}) + a_10 L ({\rm SPECID~EQ~2})
+  + a_11 L^2 ({\rm SPECID~EQ~1}) + a_12 L^2 ({\rm SPECID~EQ~2}) \right] $$
+$$ X \equiv {\rm XFOCAL} $$
+$$ Y \equiv {\rm YFOCAL} $$
+$$ L \equiv 1 - (5100\AA / \lambda)^2 $$
+
+All stellar and galaxy objects are used to fit the flux distortion
+image.  Sky fibers and QSOs are ignored in the fit.
+These objects are required to have positive flux in gri photometric imaging,
+with at least $90\%$ of the pixels good in the spectra.
+For stars, we use the PSF fluxes as measured by PHOTO.
+For galaxies, we rescale the fiber fluxes as measured by PHOTO,
+rescaled by one multiplicative factor in each band such that the
+PSF and fiber fluxes agree for stars.  Typically, that mupliplies
+the PHOTO fiber fluxes by factors of 1.37 in any band.
+
+The typical flux distortion corrections amount to ???.
 

### 2006-03-29 doc/paper/skysubtraction.tex
Update - make it tex.

@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Definitions
+%\pagestyle{empty}
+\documentclass[12pt,preprint]{aastex}
+%\documentstyle[emulateapj,danonecolfloat]{article}
+%\usepackage{emulateapj,danonecolfloat}
+\usepackage{rotating}
+
+\begin{document}
+
 \subsection{Extraction}
@@
 and second spatial derivatives, respectively).  The compiled routine
-we implement to fit the profiles in a single row (extract_row.c) can
+we implement to fit the profiles in a single row (extract\_row.c) can
 be invoked with any combination of these parameters.
@@
 3x3 polynomial distortion kernel over the entire extent of the image
-(match_trace.pro).
+(match\_trace.pro).
 
@@
 The arc-lamp reductions are written to disk at this stage, and
-stored in an extended FITS file:  HDU #0 contains the processed
-and extracted arc-lamp spectra as counts per pixel, HDU #1 contains
+stored in an extended FITS file:  HDU \#0 contains the processed
+and extracted arc-lamp spectra as counts per pixel, HDU \#1 contains
 the identified arc line wavelengths and the measured line centroids
-(including replaced positions of rejections), HDU #2 contains a single
+(including replaced positions of rejections), HDU \#2 contains a single
 structure with the Legendre polynomial coefficients applicable to each
-of the 320 arc-lamp spectra, HDU #3 contains the maskbits for each spectrum,
-and HDU #4 contains the polynomial coefficients to describe the arc-lamp
+of the 320 arc-lamp spectra, HDU \#3 contains the maskbits for each spectrum,
+and HDU \#4 contains the polynomial coefficients to describe the arc-lamp
 Gaussian widths as a function of spectral pixel.
@@
 \item{Group and sort the sky spectra as a function of wavelength.  This step
-transforms a 2-dimensional set of N_sky x 2048 pixels into a sorted 1-dimensional array.  
+transforms a 2-dimensional set of $N_{\rm sky}$ x 2048 pixels into a sorted 1-dimensional array.  
 Each pixel maintains its associated central wavelength and inverse variance. }
@@
 where inverse the 67th percentile of $\chi^2$ is greater than 1, we rescale 
-all the inverse variance values down by the 67th percentile value of $\chi^$.
+all the inverse variance values down by the 67th percentile value of $\chi^2$.
 The rescaling is actually done with a smooth interpolation as a function of 
@@
 and re-scaled errors is called 3 times successively.  After the 1st call,
-any sky fibers which have a median \chi^2 per pixel greater than 2 is 
+any sky fibers which have a median $\chi^2$ per pixel greater than 2 is 
 masked as a 'BADSKYFIBER' and the corresponding bit is set in the fibermask
@@
 width closely is closely represented by the reported variance array.
+
+\end{document}

### 2006-03-29 doc/paper/paper2d.tex
Added spectrograph info

@@
 %------------------------------------------------------------------------------
+\section{SDSS SPECTROGRAPHS}
+\label{sec_spectrographs}
+
+The SDSS spectroscopic survey was carried out with twin multi-fiber 
+spectrographs mounted at the back of the SDSS telescope (see Gunn et al.
+2006).  One of nine spectrograph cartidges can be mounted for a single
+spectroscopic observing sequence.  The cartridge holds a single aluminum
+plate, 813mm in diameter, pre-drilled and plugged with 640 fibers 
+fed into the twin spectrographs, 
+and 11 coherent fibers fed into a simple air-cooled camera (model \#??)
+for telescope guiding.
+
+The fibers are 180 $\mu m$ in diameter and subtend 2.97" at Cassegrain focus.
+The fibers accept the f/5 beam directly and, due to focal degradation, input
+an f/4 exit beam into each of the spectrographs.  The fibers are mounted in
+bundles of 20 with inter-fiber spacings of 390 $\mu m$.  Sixteen bundles are
+mounted on a single slithead with a radius of curvature of 640 mm and have
+an extra spacing of approximately 200 $\mu m$ between bundles.
+
+In Figure 1, we show a top-view schematic of one the spectrographs, and
+will detail the light path below starting at the slithead.
+The output light from the fiber is collimated with a spherical mirror
+(radius of curvature is 1264 mm), which is aligned by 3 motor mounts to
+control focus, tip \& tilt.  A left/right hartmann mask 
+covers the collimator to act as both a spectrograph shutter and traditional 
+shack-hartmann mask for focus tests.  The collimator in spectrograph 2
+was coated with an enhanced silver coating and delivers slightly higher
+reflectivity compared to the aluminum coated collimator in spectrograph 1.
+The collimated beam encounters a 40mm thick dichroic (beamsplitter) which
+reflect light blueward of 6000$\AA$.  The blue reflected light is dispersed
+with a 640 l/mm grism with zero deviation at 4960$\AA$.  The red transmitted
+light is dispersed with a 440 l/mm grism with zero deviation at 7400$\AA$.
+The dispersed light in each arm of the spectrograph enters a transmission camera
+wth a 240mm focal length (demagnification of 2.5) which is tuned for the 
+corresponding range of wavelengths.  Each of the 4 camera produce RMS 
+spot sizes of approximately 20$\mu m$ over a 16.5$\degree$ diameter 
+f.o.v. delivered to a single thinned Tektronix 2048x2048x24$\mu m$ CCD.
+An external focus ring is adjustable by hand for each CCD camera to accomodate
+differential focus changes not corrected with the common collimator.
+The four spectrograph cameras are labeled "b1, r1, b2, r2".
+
 \section{OPTIMAL EXTRACTION}

### 2006-03-30 doc/paper/paper2d.tex
B-spline updates

@@
 \label{sec_extract}
+
+We utilize an optimal extraction procedure to transform a 2-dimensional image of
+the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
+We use the same procedure on all processed frames, 
+and will describe the details in processing the flat-field frames, the
+arc lamp frames, and finally the science frames.  
+The extraction is performed row-by-row, so each frame requires an 
+outside loop over 2048 individual rows.
+The counts in each row of 2048 columns are modeled with a linear 
+combination of 320 profiles plus a low-order
+polynomial background.  In addition, deviations to the trace centers and
+the profile widths are calculated as linear basis modes (reflecting the first
+and second spatial derivatives, respectively).  The compiled routine
+we implement to fit the profiles in a single row (extract\_row.c) can
+be invoked with any combination of these parameters.
+
+We decided to perform profile fitting to the 
+SDSS spectral images for the following reasons:
+
+\begin{enumerate}
+
+\item{The fiber profiles overlap significantly with their nearest neighbors.}
+
+\item{The profiles are relatively stable and smoothly vary over 
+location on the CCDs.  They result from the convolution of the fiber 
+illumination pattern + the point-spread function
+of the optics in the SDSS spectrographs + the window funciton of the 
+Site CCD 24 micron square pixels. }
+
+\item{A least-squares fit with the true profile delivers an 
+unbiased and minimum variance estimate of the counts (e.g. Horne et al. 1995).}
+
+\item{A model profile fit leads to a model image of the SDSS data,
+and iteration allows cosmic-ray rejection and goodness-of-fit tests.}
+
+\item{A valid estimate of the error in flux measurement at each spectral pixel.}
+
+\end{enumerate}
+
+The major obstacle to actual optimal extraction is the determination of 
+the exact shape and position of the underlying profile.  
+We investigated a range of symmetric, analytic profiles with a 
+minimum number of free parameters.  In the end, we found the best 
+results using a simple Gaussian or a slightly steeper profile 
+characterized by $e^{-|x|^3/\sigma^3}$.  The profiles are defined to be 
+normalized to unity, so the amplitude of the profile fit is a 
+direct estimator of the counts at a single spectral position.
+
+Extraction is first carried out on all flat-field exposures 
+(typically 10 second observations of 4 quartz lamps projected on petals covering the entrance aperture of the telescope) which were observed
+with a common plug plate configuration (a spectroscopic sequence). 
+The flat-field extraction is used to determine the shape of
+the profile as a function of wavelength and position on the CCD.  
+The resulting spectral extractions are also used to 
+determine the empirical fiber-to-fiber variations in throughput 
+as a function of wavelength.   
+A flat-field frame is rejected by the pipeline if one of the
+following conditions is met:  1) more the 2\% of the CCD pixels are labeled
+bad by sdssproc, 2) more than 100 rows of the image have more than 20 saturated
+pixels, 
+3) the 80th percentile count level is less than 1000 counts, 
+4) fewer than all 8 flat-field telescope petals were closed, 
+5) fewer than all 4 flat-field lamps were on,
+6) more than 10000 pixels were rejected during fiber tracing, 
+7) fiber tracing fails, or traces are separated by less than 3 pixels anywhere
+on the flat-field image.
+
+The first very critical step is to robustly determine the fiber positions 
+in CCD pixel coordinates.
+This is performed by identifying peaks in a cross-sectional cut of across 
+the center row (actually a median cross-section of the central 9 rows).  
+The correct identification of the 320 fibers per frame can be 
+uniquely determined by demanding a wider spacing between the bundles 
+of 20 fibers (spaced by about 10 pixels) relative to the nominal inter-fiber 
+spacing (6.1 pixels on average).  Once the starting peaks and 
+identification have been determined, we trace the fiber centers 
+by performing flux-weighted centroids on subsequent rows
+using the previous row's center as the intial guess.  This has proven to 
+work well on SDSS flat-fields, and we have yet to uncover a failure 
+mode in over 10000 flat-field frames.
+The flux-weighted centroids do have scatter from shot noise, 
+so we fit a 7th order polynomial to each set of fiber centroids, 
+and the best-fit coefficients are stored to represent the fiber positions 
+on each flat-field frame.
+
+Each flat-field frame is extracted twice, once assuming a Gaussian
+profile and again assuming the steeper $|x|^3$ profile.  In both cases,
+the width is allowed to vary freely.  The total number of free parameters
+per row in the flat-field extractions is equal to 2 per fiber plus 10
+parameters to represent a smooth polynomial background.  After the extractions
+are complete, $\chi^2$ is computed by directly comparing the model fit
+to the flat-field image.  The median $\chi^2$ for each case is compared
+directly, and the better fit and profile (lower median $\chi^2$) are retained
+and implemented for the extraction of the remaining frames.  Also retained
+is a smooth polynomial fit the average profile width in each fiber bundle
+as a function of CCD row number.
+
+After all flat-field frames for a given sequence have been extracted
+and stored, the arc-lamp frames in the same sequence are extracted.  
+Each arc-lamp frame is matched with the nearest (in time) 
+non-rejected flat-field frame.  The matched flat-field gives the 
+initial trace centers to
+be assumed, as well as the profile shape and width determined from the
+flat-field.  To account for any flexure between the frames, the trace
+centers are allowed to shift/stretch to the arc-lamp image given a 
+3x3 polynomial distortion kernel over the entire extent of the image
+(match\_trace.pro).
+
+The arc-lamp frames are extracted row-by-row, in the same
+manner as the flat-fields, the number of free parameters assumed
+for the model fit is 2 per fiber (profile area and width) and 1 parameter
+to fit for a constant background per row.  The restricted freedom in the
+background parameterization reflects the lower scattered light levels in
+the images of the arc-lamps.
+
+\subsection{Wavelength solution}
+
+We determine an initial wavelength solution for every arc-lamp frame
+successfully extracted in the following manner:
+
+For the appropriate side of each spectrograph (red or blue), we 
+construct a list of suitable unblended lines from a parent list including
+known lines of Mercury, Cadmium, Neon and (a trace of) Argon 
+that are readily observable in the standard set of SDSS arc-lamps. 
+A representative spectrum is calculated as the median extracted spectrum
+of the 5 closest "unflagged" spectra falling near the central columns 
+of the arc-lamp image.  This median arc-spectrum is matched with 
+cross-correlation to a simulated spectrum of suitable arc-line positions and 
+strengths by systematically searching a small volume of 
+parameter space which conservatively covers
+the polynomial coefficients of the SDSS wavelength solution.
+In the first iteration, we search a set of 
+100 coefficient vectors allowing for quadratic changes and return 
+the best match.  In the second iteration, we search in a smaller region
+about the previous best match, and search a list of 125 coefficient vector
+allowing for cubic deviations.  The coefficient vector with the best
+match from the second iteration is returned as an initial guess along with
+the cross-correlation coefficient.
+
+We apply the best guess coefficients to predict the approximate pixel 
+positions of the suitable list of arc lines.  The arc line centroids
+are traced across extracted spectra (along the axis of 320 fibers).
+The centroids are fit to a smooth polynomial as a function of fiber
+position on the arc-lamp image, and subsequently fit through 6 iterations
+with simple flux weighting to arrive at a converged measure of line 
+centroids in each extracted spectrum.  Lines are excluded from the resulting
+list if the flux-weighted centroids are bad in more than 10\% of the
+spectra or if the RMS residuals are larger than 0.2 pixels.
+We perform an iterative 4th order Legendre polynomial fit of 
+logarithmic wavelength as a function of pixel in each spectrum, 
+and retain a list of rejected centroids, 
+which have deviations relative to the model of
+10\% of a pixel width.  An additional requirement is applied to each
+suitable arc line, wherein the line is rejected in no more than 3 fibers
+per bundle of 20.  Rejected centroids are replaced with the appropriate
+positions based on a linear interpolation of the differences of non-masked
+centroids relative to updated polynomial fits per spectrum.
+The final set of positions, where the same number of line positions have
+reliable centroids per spectrum, are fit with equal weighting per line
+position to produce the final wavelength
+coefficients which give the logarithmic (base 10) wavelengths as a function
+of spectral pixel.
+
+An arc-lamp image is declared unacceptable using the first 4 tests as employed
+in tests of the flat-field images.  In addition, an arc-lamp exposure
+is declared unacceptable ("BAD") if any of these following criteria are met:
+5) all 8 arc-lamps must be on, 4 Neon lamps and 4 Mercury-Cadmium lamps,
+6) the initial cross-correlation coefficient is less than 0.5,
+7) fewer than 6 arc lines are acceptable after quality assurance checks.
+
+The width of unblended arc lines are determined as an average over
+each bundle of 20 spectra.  In short, the arc-lamp emission lines 
+are extracted (with a Gaussian spectral profile) 
+allowing free parameters for the area and width of each suitable line.
+The resulting Gaussian widths per fiber bundle are fit 
+with a low order polynomial as a function of average line position
+(the fit is quadratic and cubic for the blue and red side respectively).
+The fitted quantity is referred to as the arc-lamp dispersion,
+but the values reflect the standard deviation of a Gaussian 
+profile in pixel units and represents the line-spread function of the 
+arc-lamp spectra.  Detailed analysis of the measured arc-line widths
+compared to the observed sky emission line shapes show less than a 1\%
+difference on average (McDonald et al. 2004).
+
+%%(or just refer to data model appendix instead of the paragraph below:)
+%%
+%%The arc-lamp reductions are written to disk at this stage, and
+%%stored in an extended FITS file:  HDU \#0 contains the processed
+%%and extracted arc-lamp spectra as counts per pixel, HDU \#1 contains
+%%the identified arc line wavelengths and the measured line centroids
+%%(including replaced positions of rejections), HDU \#2 contains a single
+%%structure with the Legendre polynomial coefficients applicable to each
+%%of the 320 arc-lamp spectra, HDU \#3 contains the maskbits for each spectrum,
+%%and HDU \#4 contains the polynomial coefficients to describe the arc-lamp
+%%Gaussian widths as a function of spectral pixel.
+
+\subsection{Constructing fiber flats}
+
+The final step in the reduction of the calibration frames in a given
+sequence is the construction of the fiber flat vectors.  Each flat-field
+frame is matched to the nearest arc-lamp frame as before, and the
+final flat-field fiber spectra are extracted in two steps.  The first
+step uses the preferred spatial profile and width determined above,
+and the extraction procedure produces a model of the flat-field frame.
+The model is used to subtract a diffuse scattered background given
+by an exponential point-spread function of :  (CCD halo PSF).
+We apply a 1-dimensional version of this 2-d PSF as an approximation,
+and allow the PSF to vary appropriately with wavelength.
+The final extraction is applied to the scattered light subtracted flat-field
+image allowing for 2 free paramters per fiber per row and 
+5 background parameters per row.
+
+The final flat-field spectra are stored as 
+detected counts per fiber per pixel with wavelengths of each pixel
+given by the associated solution of the arc-lamp frame.  
+We model the flat-field spectra as the product:
+
+\begin{equation} 
+{\rm Flat Model}_{ij} = A * f_{avg}(\lambda) \, {\rm fflat_i} (\lambda) 
+\end{equation}
+
+where $A$ is a overall normalization in counts, 
+$f_{avg}$ is the average response per constant logarithmic wavelength
+interval, $fflat_i$ is the fiber-to-fiber response as a function of wavelength.
+This model of the extracted flat-field spectra provides a transformation to
+compare detected counts from one fiber to another.
+
+This should remove the differential response due to the individual fiber
+throughput, as well as the small relative differences in the size of the
+spectral pixels at a constant wavelength.  ??Why do we correct for pixel
+sizes later on if the differential wavelength dispersion is included in fflat??
+
+The common normalized response, $f_{avg}$, is specified as a B-spline
+function with breakpoints spaced approximately every native pixels.
+An example of normalized individual fiber spectra 
+and the resulting B-spline fit are shown in Figure ?.  The individual fiber
+flat vectors are fits to the ratio of the extracted spectra to the superflat
+(define above).  The individual vector fits are also B-splines with breakpoints
+spaced approximately every 5 native pixels.  The final set of flat vectors are
+stored as an image with dimensions Nfiber x Npixels.
+
+\subsection{Object Extraction}
+
+With the extracted calibration frames for a given spectroscopic sequence
+reduced, we proceed to extract the remaining science frames.
+%To minimize the differences between individual science frames
+We first select a set of arc-lamp and flat-field frames as the common
+choice of initial calibrations for all science frame extractions in a given
+observing sequence.  We loop through the science exposures, processing the
+raw images with bias and pixel-to-pixel flat images (see section ??).
+We reject a science exposure outright if 1) more than 10\% of the pixels
+are flagged as bad, 2) more than 100 rows are have more than 20 saturated
+pixels, 3) the 25th
+percentile count level is greater than 1000 counts (the nominal level
+is of order 10 counts), 4) any of the flat-field petals are closed, or 5)
+any of the calibration lamps are recorded as on.
+
+Inputs to object extraction include the wavelength coefficients derived
+from the associated arc-lamp exposure, both the superflat and fiber flat
+vectors from the selected flat-field exposure, the profile shape and width
+as a function of position on the CCD image.
+
+The first step in object extraction is a simple tophat aperture (boxcar) 
+extraction centered at the fiber positions determined from the flat-field 
+and with a full spatial width of 6 CCD pixels.  The median counts per fiber 
+are passed directly to a test for overly bright fibers. Fiber spectra are 
+identified as bright (a.k.a. "WHOPPING") if the median counts per pixel in the 
+boxcar extraction are greater than 10000 counts above the median in the 
+frame.  In addition, only the brightest fiber in a spatially contiguous set of 
+bright fibers is labelled "WHOPPING".
+
+A pixelmask vector is created for each fiber spectrum to be extracted.
+See Appendix ?? for a description of the mask bits and their use.
+
+We fit for a slight distortion of the stored flat-field fiber traces to 
+best match the fiber positions of the science image.  Nominal shifts and
+distortions are usually less than a native CCD pixel.  The science frame
+is now extracted in a series of 6 steps which we detail below.
+
+\begin{enumerate}
+\item{Initial extraction including broad wings of bright (WHOPPING) fibers.}
+\item{Reconstruct a smooth background light model based on fitted parameters.}
+\item{Calculate a scattered halo image using the model of the fiber spectra less
+the aforementioned smooth background.}
+\item{Re-extract fiber spectra on the scattered halo subtracted image.}
+\item{Reconstruct smooth background and subtract.}
+\item{Final extraction with no free parameters in the background and
+no free parameters in the profile widths.}
+\end{enumerate}
+
+The extracted science spectra are first normalized by the fiber flat vectors.
+Strong atmospheric emission lines (sky lines) are identified in the the 
+set of spectra, and the wavelength solution is shifted and stretched 
+to match the identified sky line positions.  
+Examples of the shift corrections are shown in Figure ??.  
+The widths of the identified sky lines are also measured with the same method
+as the arc-line width measurements.  All spectra are also normalized by
+the average flat-field spectrum (the superflat) to remove small scale
+features likely attributed to the chromatic response of the dichroic.
+The wavelength solution is corrected to both vacuum wavelengths and the
+heliocentric velocity frame.
+
+\subsection{Sky Subtraction}
+
+Sky subtraction is performed on a frame by frame basis using the 
+"flattened" wavelength calibrated extractions.  The goal of sky subtraction
+is to estimate the expected background counts contributing to each of 
+the native 320x2048 pixels in each extracted frame.  A "supersky" model 
+at a fiducial airmass of 1 is constructed by performing iterative fits to
+the extracted spectra associated with blank sky positions (so-called
+"sky" fibers).   The following steps are taken in 3 sequential iterations
+in order to produce the sky-subtracted spectra:
+
+\begin{enumerate}
+
+\item{Identify the "sky" fibers which have no bits flagged 
+in the fiber mask array.}
+
+\item{Divide each sky spectrum by the airmass of the fiber position appropriate
+for the midpoint of the observation to produce respresentative sky spectra for 
+an airmass of 1.  The inverse variance is scaled accordingly.}
+
+\item{Group and sort the sky spectra as a function of wavelength.  This step
+transforms a 2-dimensional set of $N_{\rm sky}$ x 2048 pixels into a sorted 1-dimensional array.  
+Each pixel maintains its associated central wavelength and inverse variance. }
+
+\item{Perform a single cubic b-spline fit as a function of wavelength.  
+The number of breakpoints (number of free parameters is 4 greater) is set 
+at 2048 on the red side and 1536 on the blue side.  The breakpoint spacing 
+is set automatically to maintain approximately constant S/N between breakpoint 
+positions. The b-spline fit itself is iterative, with upper and lower rejection 
+thresholds set to mask bad or deviant pixels. }
+
+\item{The b-spline fit is evaluated at all 320x2048 wavelengths representing all
+spectral pixels in the frame.  The airmass correction is applied for each individual 
+fiber, and the rescaled super-sky is subtracted from each spectrum.}
+
+\item{Estimate the significance of the residuals in the sky fibers used for the
+fit in wavelength bins separated by breakpoints determined above.  For each bin
+where inverse the 67th percentile of $\chi^2$ is greater than 1, we rescale 
+all the inverse variance values down by the 67th percentile value of $\chi^2$.
+The rescaling is actually done with a smooth interpolation as a function of 
+redshift with a constraint that the inverse variances cannot increase. }
+
+\end{enumerate}
+
+%The b-spline routines from the numerical fortran library: {\bf slatec}
+%were ported to IDL to be easily incorporated into the pipeline.
+
+The sequence followed in the determination of the sky subtracted spectrum
+and re-scaled errors is called 3 times successively.  After the 1st call,
+any sky fibers which have a median $\chi^2$ per pixel greater than 2 is 
+masked as a 'BADSKYFIBER' and the corresponding bit is set in the fibermask
+array.  If any fibers are masked, the sky subtraction is iterated a 2nd time.
+The sky subtraction routine is invoked for one final iteration, but with extra 
+freedom allowed in the fit.  The same freedom is allowed in wavelength, by
+keeping the same number of breakpoints, but a quadratic polynomial fit is
+allowed as a function of wavelength.  This triples the number of free 
+parameters in the linear least-squares fit, but the values are still 
+overconstrained due to the ample number of sky fibers.  The extra freedom
+in the model can accomodate variation in the PSF of sky features as a function 
+of position on the CCD.  In principle, one could make use of the known 
+dispersion per spectral pixel to better model the sky spectrum
+in each fiber.
+
+The sky subtraction method which utilizes the b-spline model takes advantage 
+of the inherent over-sampling in wavelength due to the large number of 
+sky fibers assigned per plate, without the requirement of resampling the 
+pixel value to a fixed linear wavelength scale.  The least squares fit
+is carried out for native spectral pixels where the errors are very nearly 
+uncorrelated and the fit is evaluated for all the native pixels (320x2048)  
+at the central wavelengths of the single spectral frames.
+
+The accuracy of the SDSS spectral sky subtraction has been tested in recent
+scientific analyses using large numbers of background-limited spectra.   
+The distribution of flux residuals between spectral observations 
+on separate nights has been shown to be 8\% larger in width than the average expected
+from the reported errors (McDonald et al. 2004, Burgess \& Burles 2006).
+The spectra of faint, background-limited galaxies observed with SDSS exhibit a 
+distribution of residuals about the best fit spectral models described by the 
+sum of a
+Gaussian and Laplace distribution (Bolton et al. 2004).  
+The influence of the exponential tail is greatly enhanced in the vicinity of strong sky 
+emission lines representative of systematic errors in the brightest lines, but the gaussian 
+width closely is closely represented by the reported variance array.
 
@@
 
+\bibitem[de Boor 1978]{deboor78}
+de Boor, C. 1978, A Practical Guide to Splines, Springer, New York.
+
 %Gray's code SPECTRUM is
@@
 
-What the heck is a B-spline?
+In many of our algorithms we utilize a class of fitting functions known
+as B-splines (de Boor 1978).  These are very flexible and efficient 
+continous functions which have been applied to numerous spectroscopic
+and photometric data sets (cite Kelson ??, Kirkman/O'Meara ??, 
+Bolton et al. 2005, Davis/Newman/Finkbeiner).  We also extend the 1-dimensional
+B-splines to slightly higher dimension by allowing the B-spline evalutions
+represent polynomial coefficients in a second independent variable.  We 
+call these 2-dimensional B-splines, but in fact, 
+they are a restricted subset of the full 2-dimensional multi-variate B-splines 
+that allow the same degree of freedom in both independent variable domains.
+
+In its basic form, a B-spline represents
+a continous set of piecewise polynomials as a function of a single independent
+variable.  The b-spline of order $n$ is equivalent to a set 
+of individual polynomial curves (order 4 = cubic, for example) with $m$ 
+breakpoints at positions of the boundaries of the individual polynomial 
+curves in the independent variable domain.  Unconstrained, the fitting function
+would have a total number of free parameters = $n * (m-1)$.  The B-spline is
+constrained such that only the highest non-trivial spatial derivative is
+allowed to change, which sums to a total number of constraints = 
+$(n-1) * (m-2)$.  A B-spline of order $n$ is therefore described by 
+$n + m - 2$ free parameters.  A simple example is to set m=2 breakpoints at
+the extreme range of the independent variable domain, and the B-spline 
+represents a simple single polynomial function with $n$ free parameters.
+Cast in this form, each B-spline coefficient is correlated with the
+$n-1$ nearest neighbors.  This allows for very efficient banded linear matrix
+inversion.  For our purposes, we have adopted the original fortran subroutines
+in the slatec library (ref??) and translated completely to IDL.
 

### 2006-03-30 doc/paper/paper2d.tex
Wednesday night update

@@
 
-\item{Perform a single cubic b-spline fit as a function of wavelength.  
+\item{Perform a single cubic B-spline fit as a function of wavelength.  
 The number of breakpoints (number of free parameters is 4 greater) is set 
@@
 is set automatically to maintain approximately constant S/N between breakpoint 
-positions. The b-spline fit itself is iterative, with upper and lower rejection 
+positions. The B-spline fit itself is iterative, with upper and lower rejection 
 thresholds set to mask bad or deviant pixels. }
 
-\item{The b-spline fit is evaluated at all 320x2048 wavelengths representing all
+\item{The B-spline fit is evaluated at all 320x2048 wavelengths representing all
 spectral pixels in the frame.  The airmass correction is applied for each individual 
@@
 
-%The b-spline routines from the numerical fortran library: {\bf slatec}
+%The B-spline routines from the numerical fortran library: {\bf slatec}
 %were ported to IDL to be easily incorporated into the pipeline.
@@
 
-The sky subtraction method which utilizes the b-spline model takes advantage 
+The sky subtraction method which utilizes the B-spline model takes advantage 
 of the inherent over-sampling in wavelength due to the large number of 
@@
 The set of all blue and red exposures for each fiber are
-combined using a B-spline.
+combined using an iterative B-spline fit to reject outlying flux pixels
+and remaining cosmic rays.
 The splines are evaluated on a wavelength grid that
@@
 
+\bibitem[Bolton et al.(2006)]{2006ApJ...638..703B} Bolton, A.~S., Burles, 
+S., Koopmans, L.~V.~E., Treu, T., \& Moustakas, L.~A.\ 2006, \apj, 638, 703
+
+\bibitem[Bolton et al.(2004)]{2004AJ....127.1860B} Bolton, A.~S., Burles, 
+S., Schlegel, D.~J., Eisenstein, D.~J., \& Brinkmann, J.\ 2004, \aj, 127, 
+1860
+
+\bibitem[de Boor 1977]{deboor77}
+de Boor, C. 1977,  SIAM Journal on Numerical Analysis 14, 3, 441
+
 \bibitem[de Boor 1978]{deboor78}
@@
 
+\bibitem[Gunn et al.(2006)]{2006astro.ph..2326G} Gunn, J.~E., Siegmund, 
+W.~A., \& Mannery, E.~J.\ 2006, arXiv:astro-ph/0602326
+
+\bibitem[Kelson(2003)]{2003PASP..115..688K} Kelson, D.~D.\ 2003, \pasp, 
+115, 688
+
+\bibitem[Nollett \& Burles(2000)]{2000PhRvD..61l3505N} Nollett, K.~M., \& 
+Burles, S.\ 2000, \prd, 61, 123505
+
 \bibitem[O'Donnell 1994]{odonnell94} 
@@
 Schlegel, D.\ J., Finkbeiner, D.\ P., \& Davis M.\ 1998, \apj, 500, 525 (SFD)
+
+\bibitem[Tytler et al.(2004)]{2004ApJ...617....1T} Tytler, D., et al.\ 
+2004, \apj, 617, 1
+
+\bibitem[Uomoto et al.(1999)]{1999AAS...195.8701U} Uomoto, A., et al.\ 
+1999, Bulletin of the American Astronomical Society, 31, 1501
 
@@
 continous functions which have been applied to numerous spectroscopic
-and photometric data sets (cite Kelson ??, Kirkman/O'Meara ??, 
-Bolton et al. 2005, Davis/Newman/Finkbeiner).  We also extend the 1-dimensional
+and photometric data sets (cite Kelson ??, Tytler et al. 2004, 
+Bolton et al. 2005, Davis/Newman/Finkbeiner) as well as other interpolative
+analyses (e.g. Nollett \& Burles 2000).  
+We also extend the 1-dimensional
 B-splines to slightly higher dimension by allowing the B-spline evalutions
@@
 a continous set of piecewise polynomials as a function of a single independent
-variable.  The b-spline of order $n$ is equivalent to a set 
+variable.  The B-spline of order $n$ is equivalent to a set 
 of individual polynomial curves (order 4 = cubic, for example) with $m$ 
@@
 $n + m - 2$ free parameters.  A simple example is to set m=2 breakpoints at
-the extreme range of the independent variable domain, and the B-spline 
+the extreme range of the independent variable domain, and the B-spline then
 represents a simple single polynomial function with $n$ free parameters.
@@
 inversion.  For our purposes, we have adopted the original fortran subroutines
-in the slatec library (ref??) and translated completely to IDL.
+in the slatec library (de Boor 1977) to complete translation in IDL.
 

### 2006-03-30 doc/paper/paper2d.tex
Update.

@@
 The wavelength calibration is shown to have an accuracy of $3 \kms$,
-and the broad-band spectrophotometry has an accuracy of better than $10\%$.
+and the broad-band spectrophotometry has an accuracy of better than $6\%$.
 
@@
 the PHOTO fiber fluxes by factors of 1.37 in any band.
+Both PSF fluxes and fiber fluxes are converted to AB magnitudes
+using the following:
+$$ u_{AB} = u - 0.042 $$
+$$ g_{AB} = g + 0.036 $$
+$$ r_{AB} = r + 0.015 $$
+$$ i_{AB} = i + 0.013 $$
+$$ z_{AB} = z - 0.002 $$
+{\bf See D. Hogg sdss-calib/845}
 

### 2006-03-31 doc/paper/paper2d.tex
Update.

@@
 The wavelength solution is corrected to both vacuum wavelengths and the
-heliocentric velocity frame.
+barycenter velocity frame.
 
@@
 width closely is closely represented by the reported variance array.
+
+\section{Testing velocities}
+
+% See sdss-spcommish/206 for description of plate 321
+
+A plate was drilled centered on the open cluster M67 with the
+specific idea of testing the SDSS velocity precision.
+130 stars were targetted from the radial velocity catalog
+of Mathieu \etal\ (1986), falling in the magnitude range of 8 to 14.
+The 19 stars brighter than $m=11$ were not plugged, since they
+would have saturated the detectors.  This was plate 321 MJD 51612.
 
@@
 
+% Precise radial velocities of late-type stars in the open clusters M11 and M67
+\bibitem[Mathieu \etal\ (1896)]{1986AJ...92...1100} Mathieu, R.~D.,
+Latham, D.~W., Griffin, R.~F., \& Gunn, J.~E.\ 1986, \aj, 92, 1100
+
 \bibitem[Nollett \& Burles(2000)]{2000PhRvD..61l3505N} Nollett, K.~M., \& 
@@
 %------------------------------------------------------------------------------
-\appendix
 \section{DATA MODEL}
-\label{bsplines}
-
-What the heck is a B-spline?
+\label{datamodel}
+
+This spCFrame files contain the extracted spectra for a single CCD of a
+single exposure. In general, each object has several blue exposures and
+several red exposures. These spectra are still in the native wavelength
+mapping, which is neither linear in wavelength nor log-wavelength.
+However, the spectra are flux-calibrated, and deviant pixels are
+rejected in the pixel mask (from the comparison with the other exposures).
+The spCFrame file contains the following:
+\begin{verbatim}
+HDU #0 = Flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
+HDU #1 = Inverse variance (1/sigma^2) for the above [FLOAT]
+HDU #2 = Pixel mask [32-bit INT]
+HDU #3 = Wavelength solution in log-10 Angstroms (vacuum) [FLOAT]
+HDU #4 = Wavelength dispersion in units of (10e-4 log-10 wavelength)^(-1) [FLOAT]
+         (these are the same units as in the combined-spectrum spPlate file)
+HDU #5 = Plug-map structure from plPlugMapM file [BINARY FITS TABLE]
+HDU #6 = Sky flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
+\end{verbatim}
+
+The spPlate files contains the combined spectra for all exposures
+(potentially spanning multiple nights) for a given mapped plate.
+The first 5 HDU's contain NPIX by 640 images, where NPIX is the number
+of wavelengths. The pixel scale is exactly 1e-4 in log-wavelength
+(any units), which corresponds to $10^{-4} c \ln(10) = 69.029765 \kms$.
+The spPlate file contains the following:
+\begin{verbatim}
+HDU #0 = Flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
+HDU #1 = Inverse variance (1/sigma^2) for the above [FLOAT]
+HDU #2 = AND mask [32-bit INT]
+HDU #3 = OR mask [32-fit INT]
+HDU #4 = Wavelength dispersion in pixels [FLOAT]
+HDU #5 = Plug-map structure from plPlugMapM file [BINARY FITS TABLE]
+HDU #6 = Average sky flux in units of 10^(-17) erg/s/cm^2/Ang [FLOAT]
+\end{verbatim}
+
+The following header keywords are of interest (in HDU \#0):
+\begin{verbatim}
+PLATEID = Plate ID
+TILEID  = Tile ID
+MJD     = Modified Julian date for first night of observation
+MJDLIST = Modified Julian date for all nights of observation
+MAPID   = Fiber mapper ID
+CARTID  = Fiber cartridge used at the telescope
+NAME    = Name of target: Plate ID, MJD, Fiber mapper rerun
+COEFF0  = Central wavelength (log10-Angstroms) of first pixel
+COEFF1  = Wavelength range of each pixel (log-10 Angstroms)
+CRVAL1  = (Same as COEFF0)
+CD1_1   = (Same as COEFF1)
+RADEG   = Right ascension of plate center (J2000 degrees)
+DECDEG  = Declination of plate center (J2000 degrees)
+AIRMASS = Airmass for mid-point of observation
+RA      = Right ascension of where the telescope thinks it is pointing
+DEC     = Declination of where the telescope thinks it is pointing
+AZ      = Azimuth of the telescope (degrees)
+ALT     = Altitude of the telescope (degrees)
+NEXP    = Total number of frames from all cameras used in the co-adding;
+          since there are 4 cameras, NEXP=12 would typically mean that there
+          were 3 exposures in each camera.
+EXPT_B1 = b1 camera exposure time (seconds)
+EXPT_B2 = b2 camera exposure time (seconds)
+EXPT_R1 = r1 camera exposure time (seconds)
+EXPT_R2 = r2 camera exposure time (seconds)
+EXPTIME = Minimum of exposure times for all cameras
+SPEC1_G = (S/N)^2 per pixel at g=20.2 on spectrograph #1
+SPEC2_G = (S/N)^2 per pixel at g=20.2 on spectrograph #2
+SPEC1_R = (S/N)^2 per pixel at r=20.25 on spectrograph #1
+SPEC2_R = (S/N)^2 per pixel at r=20.25 on spectrograph #2
+SPEC1_I = (S/N)^2 per pixel at i=19.9 on spectrograph #1
+SPEC2_I = (S/N)^2 per pixel at i=19.9 on spectrograph #2
+GOFFSTD = Median spectrophotometry minus photo mag for standard stars in g-band
+ROFFSTD = Median spectrophotometry minus photo mag for standard stars in r-band
+IOFFSTD = Median spectrophotometry minus photo mag for standard stars in i-band
+GOFFRMS = RMS of spectrophotometry minus photo mag for standard stars in g-band
+ROFFRMS = RMS of spectrophotometry minus photo mag for standard stars in r-band
+IOFFRMS = RMS of spectrophotometry minus photo mag for standard stars in i-band
+GOFFGAL = Median spectrophotometry minus photo mag for "main" galaxies in g-band
+ROFFGAL = Median spectrophotometry minus photo mag for "main" galaxies in r-band
+IOFFGAL = Median spectrophotometry minus photo mag for "main" galaxies in i-band
+GOFFGAL = RMS of spectrophotometry minus photo mag for "main" galaxies in g-band
+ROFFGAL = RMS of spectrophotometry minus photo mag for "main" galaxies in r-band
+IOFFGAL = RMS of spectrophotometry minus photo mag for "main" galaxies in i-band
+BESTEXP = Exposure number for the "best" exposure (all 4 cameras)
+EXPIDxx = Identifier for the each invididual exposure+CCD used in this combine,
+          for ex. 'b1-00006853-00006857-00006858' refers to the b1 CCD,
+          exposure #6853, using flat #6857 and arc #6858
+TAI     = Number of seconds since Nov 17 1858 for mid-point of observation
+TAI-BEG = Number of seconds since Nov 17 1858 for beginning of observation
+TAI-END = Number of seconds since Nov 17 1858 for end of observation
+SEEING20= 20th-percentile of seeing during exposure (arcsec)
+SEEING50= 50th-percentile of seeing during exposure (arcsec)
+SEEING80= 80th-percentile of seeing during exposure (arcsec)
+RMSOFF20= 20th-percentile of RMS offset of guide fibers (arcsec)
+RMSOFF50= 50th-percentile of RMS offset of guide fibers (arcsec)
+RMSOFF80= 80th-percentile of RMS offset of guide fibers (arcsec)
+VERSIDL = Version of IDL
+VERS2D  = Version of idlspec2d for 2D reduction
+VERSCOMB= Version of idlspec2d for combining multiple spectra
+VERSFLAT= Version of specflat for spectroscopic flat-fields
+VERS1D  = Version of idlspec2d for 1D reduction
+HELIO_RV= Barycentric velocity correction (km/sec)
+\end{verbatim}
+
+There are two masks, an "AND" mask and an "OR" mask. The spectra
+are constructed from 3 or more 15-minute observations, and the
+"AND" mask bits are set if that bit is set for each and every
+input observation. The "OR" mask bits are set if that bit is set
+for any of the observations. Usually, I only look at the "AND" mask.
+The mask bits are set as follows:
+\begin{verbatim}
+ 0 NOPLUG          Fiber not listed in plugmap file
+ 1 BADTRACE        Bad trace from routine TRACE320CRUDE
+ 2 BADFLAT         Low counts in fiberflat
+ 3 BADARC          Bad arc solution
+ 4 MANYBADCOLUMNS  >10% pixels are bad columns
+ 5 MANYREJECTED    >10% pixels are rejected in extraction
+ 6 LARGESHIFT      Large spatial shift between flat and object pos'n
+ 7 BADSKYFIBER     Sky Fiber shows extreme residuals
+ 8 NEARWHOPPER     Within 2 fibers of a whopping fiber (inclusive)
+ 9 WHOPPER         Whopping fiber
+16 NEARBADPIXEL    Bad pixel within 3 pixels of trace
+17 LOWFLAT         Flat field less than 0.5
+18 FULLREJECT      Pixel fully rejected in extraction (INVVAR=0)
+19 PARTIALREJECT   Some pixels rejected in extraction
+20 SCATTEREDLIGHT  Scattered light significant
+21 CROSSTALK       Cross-talk significant
+22 NOSKY           Sky level unknown at this wavelength (INVVAR=0)
+23 BRIGHTSKY       Sky level > flux + 10*(flux error)
+                   AND sky > 2.0 * median(sky,99 pixels)
+24 NODATA          No data available in combine B-spline (INVVAR=0)
+25 COMBINEREJ      Rejected in combine B-spline
+26 BADFLUXFACTOR   Low flux-calibration or flux-correction factor
+27 BADSKYCHI       Relative chi^2 > 3 in sky residuals at this wavelength
+28 REDMONSTER      Contiguous region of bad chi^2 in sky residuals
+                   (with threshhold of relative chi^2 > 3)
+\end{verbatim}
+When low numbered bits ($<16$) are set, those will be set for half of the
+spectra -- either the blue or red spectrograph. The higher-numbered bits
+($>=16$) are set for individual pixels.
+
+Which mask bits are important?
+The conditions that are considered very bad are already used to set
+the errors to infinity for the effected pixels (specifically, the
+inverse variance is set to zero). The most useful mask bit to look
+at is BRIGHTSKY, which indicates when the sky is so bright relative
+to the object that perhaps one shouldn't trust any of the object
+flux there. Our reported errors are meant to include sky-subtraction
+errors, but there are instances (particularly around 5577) where
+these errors may be untrustworthy.
+
+The dispersion per pixel and the sky flux are computed at each pixel
+by re-weighting the individual spectra at each pixel according to
+their formal errors. This re-weighting is only approximate.
+
+Note that the sky lines are slightly shifted in the reductions
+because we transform the velocities to the barycenter of the solar
+system. Each exposure that contributes to the co-added spectra will
+have slightly different barycenter correction, so the "average sky"
+contains a superposition of these slightly-offset sky lines. These
+shifts keep the object spectra as-measured at the barycenter,
+regardless of the time of year or the Earth's rotation relative to
+the spectroscopic targets.
+
+\subsection{Target Flags}
+
+The target selection sets bits in two flags called PRIMTARGET and
+SECTARGET. An object can have multiple bits set. For example, if
+the target selection code determined that an object was both a
+QSO\_SKIRT and a STAR\_WHITE\_DWARF, then two bits will be set. The
+value of PRIMTARGET in this example will be $2^4 + 2^{19} = 524304$.
+
+The PRIMTARGET bit values are as follows (in hex and decimal):
+\begin{verbatim}
+  QSO_HIZ           = 0x1          = bit # 0 =        1
+  QSO_CAP           = 0x2          = bit # 1 =        2
+  QSO_SKIRT         = 0x4          = bit # 2 =        4
+  QSO_FIRST_CAP     = 0x8          = bit # 3 =        8
+  QSO_FIRST_SKIRT   = 0x10         = bit # 4 =       16
+  QSO_MAG_OUTLIER   = 0x2000000    = bit #25 = 33554432
+  QSO_REJECT        = 0x20000000   = bit #29 =536870912
+  GALAXY_RED        = 0x20         = bit # 5 =       32
+  GALAXY_RED_II     = 0x4000000    = bit #26 = 67108864
+  GALAXY            = 0x40         = bit # 6 =       64
+  GALAXY_BIG        = 0x80         = bit # 7 =      128
+  GALAXY_BRIGHT_CORE= 0x100        = bit # 8 =      256
+  ROSAT_A           = 0x200        = bit # 9 =      512
+  ROSAT_B           = 0x400        = bit #10 =     1024
+  ROSAT_C           = 0x800        = bit #11 =     2048
+  ROSAT_D           = 0x1000       = bit #12 =     4096
+  ROSAT_E           = 0x8000000    = bit #27 =134217728
+  STAR_BHB          = 0x2000       = bit #13 =     8192
+  STAR_CARBON       = 0x4000       = bit #14 =    16384
+  STAR_BROWN_DWARF  = 0x8000       = bit #15 =    32768
+  STAR_SUB_DWARF    = 0x10000      = bit #16 =    65536
+  STAR_CATY_VAR     = 0x20000      = bit #17 =   131072
+  STAR_RED_DWARF    = 0x40000      = bit #18 =   262144
+  STAR_WHITE_DWARF  = 0x80000      = bit #19 =   524288
+  STAR_PN           = 0x10000000   = bit #28 =268435456
+  SERENDIP_BLUE     = 0x100000     = bit #20 =  1048576
+  SERENDIP_FIRST    = 0x200000     = bit #21 =  2097152
+  SERENDIP_RED      = 0x400000     = bit #22 =  4194304
+  SERENDIP_DISTANT  = 0x800000     = bit #23 =  8388608
+  SERENDIP_MANUAL   = 0x100000     = bit #24 = 16777216
+  QSO_MAG_OUTLIER   = 0x200000     = bit #25 = 33554432
+  GALAXY_RED_II     = 0x400000     = bit #26 = 67108864
+  ROSAT_E           = 0x800000     = bit #27 =134217728
+  STAR_PN           = 0x1000000    = bit #28 =268435456
+  QSO_REJECT        = 0x2000000    = bit #29 =536870912
+  SOUTHERN_SURVEY   = 0x8000000    = bit #31 =-2147483648
+\end{verbatim}
+
+The SECTARGET bit values are as follows (in hex and decimal):
+\begin{verbatim}
+  LIGHT_TRAP        = 0x1          = bit # 0 =        1
+  REDDEN_STD        = 0x2          = bit # 1 =        2
+  TEST_TARGET       = 0x4          = bit # 2 =        4
+  QA                = 0x8          = bit # 3 =        8
+  SKY               = 0x10         = bit # 4 =       16
+  SPECTROPHOTO_STD  = 0x20         = bit # 5 =       32
+  GUIDE_STAR        = 0x40         = bit # 6 =       64
+  BUNDLE_HOLE       = 0x80         = bit # 7 =      128
+  QUALITY_HOLE      = 0x100        = bit # 8 =      256
+  HOT_STD           = 0x200        = bit # 9 =      512
+  SOUTHERN_SURVEY   = 0x8000000    = bit #31 =-2147483648
+\end{verbatim}
+
+\subsection{Vacuum wavelengths}
+
+The data are stored using vacuum wavelengths.
+Most optical astronomers know the wavelengths of transitions
+as measured at S.T.P., which is how the CRC lists them for any
+transitions redward of $2000\AA$.
+
+The IAU standard for conversion from air to vacuum wavelengths
+is given in Morton (1991, ApJS, 77, 119). For vacuum wavelengths ($\lambda$)
+in Angstroms, convert to air wavelength ($\lambda_{\rm air}$) via:
+$$\lambda_{\rm air} = \lambda / (1 + 2.735182E-4 + 131.4182 / \lambda^2 + 2.76249E8 / \lambda^4) $$
+
+Below, I've listed the vacuum wavelength of some transitions:
+\begin{verbatim}
+ Air       Vacuum
+ --------  --------
+ 4861.363  4862.721  H_beta
+ 4958.911  4960.295  [O III]
+ 5006.843  5008.239  [O III]
+ 6548.05   6549.86   [N II]
+ 6562.801  6564.614  H_alpha
+ 6583.45   6585.27   [N II]
+ 6716.44   6718.29   [S II]
+ 6730.82   6732.68   [S II]
+\end{verbatim}
+The wavelengths are shifted such that measured velocities are relative
+to barycentric at the mid-point of each 15-minute exposure.
 

### 2006-03-31 doc/paper/paper2d.tex
updated observations section

@@
 
+\section{Typical Spectroscopic Observations}
+\label{sec_observ}
+
+At the beginning of a typical survey night, up to 9 plates have been mounted
+in cartridges, plugged with 640 science fibers and 11 guide fibers. 
+The position of the 640 fibers along each of the two fiber slitheads
+are mapped by the 8th wonder of the world,
+the SDSS plugplate fiber mapper (Schlegel and Finkbeiner).
+All usable spectroscopic data obtained for a single plugging, 
+and therefore a single fiber mapping, will eventually be combined into 
+a final data product for that plugging.  In this scheme, an individual 
+output spectrum is uniquely specified by the plate number, the first night
+of observations (in MJD), and the science fiber number.
+
+A plate is mounted in the telescope through a efficient procedure described
+in Gunn et al. 2006.  The switch from the imaging camera to a spectroscopic
+cartridge takes an average of xx minutes, while a switch from one cartridge
+to another takes merely x minutes.  Once the cartridge is mounted, the telescope is slewed to the coordinates of the plate center.  
+Before acquisition of the guide stars, the 8 flat-field petals are closed and
+a sequence of one 10-second flat-field exposure and one 2-second 
+Neon-Mercury-Cadmium arc exposure are obtained.  While the arc-lamp frame is
+being readout, the flat-field petals are opened, and guide stars are acquired
+through the 11 coherent guide bundles (9 are this big and 2 are even bigger??).
+With at least 4 guide stars acquired, closed loop guiding is started, and
+usually converges to stable guiding within 2 minutes??
+The plate scale of the SDSS telescope can be adjusted, and the guiding software
+suggest a scale change if one is required.  Once close loop 
+guiding has settled, the first science exposure is started.  The typical 
+exposure time is 900 seconds, but exposure times from 600 to 1200 second are
+not unexpected based on conditions and time constraints.  A second exposure
+is started immediately after the first science exposure is complete and readout.A fast linux workstation onsite copies the data from the data acquisition host,
+and automatically performs a quick spectral extraction with a subset of the 
+steps explained here.  The results are posted to sos.apo.nmsu.edu, and
+the SDSS observers can check the S/N obtained in each of the 4 spectroscopic
+cameras in real-time.  A plate is considered to have met SDSS survey standards
+if all 4 cameras have a cumulative (S/N)^2 greater than 15 at 
+fidicual magnitudes of ?? and ??.
+
 \section{OPTIMAL EXTRACTION}
@@
 we implement to fit the profiles in a single row (extract\_row.c) can
-be invoked with any combination of these parameters.
+be invoked with any combination of these parameters.  The parameters are
+sorted in order of fiber number with background parameters included at the
+end of the list.  This facilitates and banded matrix inversion, and 
+significantly increases the speed of Cholesky decomposition step in each
+iteration of the extraction.
 
@@
 frame.  In addition, only the brightest fiber in a spatially contiguous set of 
-bright fibers is labelled "WHOPPING".
+bright fibers is labelled "WHOPPING".  Each whopping fiber adds an additional
+background parameter to the fit per row, which is an exponential profile
+with a spatial scale of 25 pixels.  Such a profile correlates with many
+neighboring fibers, so the parameter is treated as a background to supplement
+the smooth polynomial background.  The flux contained in the exponential 
+profile is not included in the extracted spectra.
 
@@
 
+%------------------------------------------------------------------------------
+% ACKNOWLEDGMENTS
+%------------------------------------------------------------------------------
+\bigskip
+We are very grateful to Michael Strauss for
+useful and helpful comments on the manuscript.
+Funding for the creation and distribution of the SDSS Archive has been
+provided by the Alfred P. Sloan Foundation, the Participating
+Institutions, the National Aeronautics and Space Administration, the
+National Science Foundation, the U.S. Department of Energy, the
+Japanese Monbukagakusho, and the Max Planck Society. The SDSS Web site
+is http://www.sdss.org/. The SDSS is managed by the Astrophysical
+Research Consortium (ARC) for the Participating Institutions. The
+Participating Institutions are The University of Chicago, Fermilab,
+the Institute for Advanced Study, the Japan Participation Group,
+The Johns Hopkins University, the Korean Scientist Group, Los Alamos
+National Laboratory, the Max-Planck-Institute for Astronomy (MPIA),
+the Max-Planck-Institute for Astrophysics (MPA), New Mexico State
+University, University of Ohio, University of Pittsburgh,
+University of Portsmouth,
+Princeton University, the United States Naval Observatory, and
+the University of Washington.
+
 \newpage
@@
 1999, Bulletin of the American Astronomical Society, 31, 1501
+
+\bibitem[Wild \& Hewett(2005)]{2005MNRAS.358.1083W} Wild, V., \& Hewett, 
+P.~C.\ 2005, \mnras, 358, 1083
 

### 2006-03-31 doc/paper/paper2d.tex
Make it tex.
Added disp figure caption

@@
 cameras in real-time.  A plate is considered to have met SDSS survey standards
-if all 4 cameras have a cumulative (S/N)^2 greater than 15 at 
+if all 4 cameras have a cumulative (S/N)$^2$ greater than 15 at 
 fidicual magnitudes of ?? and ??.
@@
 detected counts per fiber per pixel with wavelengths of each pixel
-given by the associated solution of the arc-lamp frame.  
+given by the solution of the matched arc-lamp frame.  
 We model the flat-field spectra as the product:
@@
 \begin{equation} 
-{\rm Flat Model}_{ij} = A * f_{avg}(\lambda) \, {\rm fflat_i} (\lambda) 
+{\rm Model}_{ij} = A \times {\rm SF}(\lambda) \times {\rm (fiber~flat)_i} (\lambda) 
 \end{equation}
@@
 where $A$ is a overall normalization in counts, 
-$f_{avg}$ is the average response per constant logarithmic wavelength
-interval, $fflat_i$ is the fiber-to-fiber response as a function of wavelength.
+$SF$ is the average wavelength dependent response called the ``superflat'', and
+${\rm fiber~flat_i}$ is the fiber-to-fiber response as a function of wavelength.
 This model of the extracted flat-field spectra provides a transformation to
@@
 Sky subtraction is performed on a frame by frame basis using the 
-"flattened" wavelength calibrated extractions.  The goal of sky subtraction
+``flattened'' wavelength calibrated extractions.  The goal of sky subtraction
 is to estimate the expected background counts contributing to each of 
-the native 320x2048 pixels in each extracted frame.  A "supersky" model 
+the native 320x2048 pixels in each extracted frame.  A ``supersky'' model 
 at a fiducial airmass of 1 is constructed by performing iterative fits to
 the extracted spectra associated with blank sky positions (so-called
-"sky" fibers).   The following steps are taken in 3 sequential iterations
+``sky'' fibers).   The following steps are taken in 3 sequential iterations
 in order to produce the sky-subtracted spectra:
@@
 is uniform is log-wavelengths, with the scale always chosen
-as $10^{-4}$ in log-wavelength.  The wavelength ranges are
+as $10^{-4}$ in log$_{10}$-wavelength (pixels are constant in 
+velocity width with $\Delta v = 69.02$ km/s).  The wavelength ranges are
 slightly different for each plate, chosen to span the input data.
@@
 and $9000\AA$.
-There are also magnitude offsets as a function fo spectrograph ID,
+There are also magnitude offsets as a function of spectrograph ID,
 and a chromatic offset as a function of spectrograph ID.
-$$ F' = F  \times \left[ a_0 ({\rm SPECID~EQ~1}) + a_1 ({\rm SPECID~EQ~2}) \right]
-  \times \exp \left[ a_2 x + a_3 y + a_4 xy + a_5 x^2 + a_6 y^2 + a_7 xL + a_8 yL
-  + a_9 L ({\rm SPECID~EQ~1}) + a_10 L ({\rm SPECID~EQ~2})
-  + a_11 L^2 ({\rm SPECID~EQ~1}) + a_12 L^2 ({\rm SPECID~EQ~2}) \right] $$
+$$ F' =  F  \times \left[ a_0 {\rm S_1} + a_1 {\rm S_2} \right]
+ \times \exp \left[ a_2 x + a_3 y + a_4 xy + a_5 x^2 + a_6 y^2  + a_7 xL + a_8 yL \right] $$
+$$  \times \exp \left[ a_9 L \, {\rm S_1} + a_{10} L \, {\rm S_2}
+  + a_{11} L^2 \, {\rm S_1} + a_{12} L^2 \, {\rm S_2} \right]  $$
+$$ S_1 \equiv {\rm SPECID~EQ~1} $$
+$$ S_2 \equiv {\rm SPECID~EQ~2} $$
 $$ X \equiv {\rm XFOCAL} $$
@@
 in Angstroms, convert to air wavelength ($\lambda_{\rm air}$) via:
-$$\lambda_{\rm air} = \lambda / (1 + 2.735182E-4 + 131.4182 / \lambda^2 + 2.76249E8 / \lambda^4) $$
+$$\lambda_{\rm air} = \lambda / (1 + 0.0002735182 + 131.4182 / \lambda^2 + 2.76249E8 / \lambda^4) $$
 
@@
 %------------------------------------------------------------------------------
+%% Figures
+
+\begin{figure}[tb]
+\figurenum{7}
+\epsscale{0.9}
+\plotone{plot_disp.ps}
+\figcaption{Measurements of spectral resolution in the four SDSS 
+spectrograph cameras.  For each camera, we should the median
+FWHM arc-line widths as a function of wavelength as 
+solid lines for cameras b1 (cyan) and r1 (orange),
+dashed lines for cameras b2 (blue) and r2 (red).
+The range of resolution is shown as error bars representing
+the 10\% to 90\% ranges in each camera over all fibers and
+all plates through DR5.
+\label{fig7}
+}
+\end{figure}
 

### 2006-04-01 doc/paper/paper2d.tex
Update.

@@
   QSO_FIRST_SKIRT   = 0x10         = bit # 4 =       16
-  QSO_MAG_OUTLIER   = 0x2000000    = bit #25 = 33554432
-  QSO_REJECT        = 0x20000000   = bit #29 =536870912
   GALAXY_RED        = 0x20         = bit # 5 =       32
-  GALAXY_RED_II     = 0x4000000    = bit #26 = 67108864
   GALAXY            = 0x40         = bit # 6 =       64
@@
   ROSAT_D           = 0x1000       = bit #12 =     4096
-  ROSAT_E           = 0x8000000    = bit #27 =134217728
   STAR_BHB          = 0x2000       = bit #13 =     8192
@@
   STAR_WHITE_DWARF  = 0x80000      = bit #19 =   524288
-  STAR_PN           = 0x10000000   = bit #28 =268435456
   SERENDIP_BLUE     = 0x100000     = bit #20 =  1048576

### 2006-04-01 doc/paper/paper2d.tex
Checking in the plate statistics plot

@@
 large scale structure of the Universe and other extragalactic science.
+
+In this paper, we describe the details of the SDSS 2-d spectroscopic pipeline
+(idlspec2d) including steps starting with processing of the raw CCD frames to final
+flux calibration and spectral co-addition.  Downstream pipelines which
+calculate redshifts, spectral types, emission and absorption line identification,
+and velocity dispersions are described elsewhere.
+As of writing the 2-d pipeline has been run successfully on all spectroscopic
+data obtained in the course of the survey from February 28, 2000 (MJD 2451602) 
+to June 15, 2005 (MJD 2453536).  Over this timespan, which includes public data 
+releases EDR through DR5, SDSS obtained survey quality spectroscopic data on
+1656 fiber pluggings (640 fibers each in 1575 unique plates).   
+A total of 972,111 sky positions were targetted, with the following 
+breakdown in the best matched spectral classification:  651920 galaxies, 
+208054 stars, 112137 QSOs and 51300 sky spectra.
+
+Mean exposure time per plate is 3914 seconds, and the median is 3302 seconds.
+In Figure 1 we show three of the observing diagnostics, on the top is
+the effective signal-to-noise of the whole plate normalized to 3600 second
+exposure time, the median seeing in arcseconds and the rms guide probe offset
+in arcseconds are shown below.  The abscissa is in days after Feb 26, 2000
+(MJD 2451600).
 

### 2006-04-01 doc/paper/paper2d.tex
example of superflat

@@
 compare detected counts from one fiber to another.
-
-This should remove the differential response due to the individual fiber
+The flat-field vectors correct the differential response due to the individual fiber
 throughput, as well as the small relative differences in the size of the
-spectral pixels at a constant wavelength.  ??Why do we correct for pixel
-sizes later on if the differential wavelength dispersion is included in fflat??
-
-The common normalized response, $f_{avg}$, is specified as a B-spline
+spectral pixels at a constant wavelength.
+
+The common normalized response, $SF$, is specified as a B-spline
 function with breakpoints spaced approximately every native pixels.
 An example of normalized individual fiber spectra 
-and the resulting B-spline fit are shown in Figure ?.  The individual fiber
+and the resulting B-spline fit are shown in Figure 2.  The individual fiber
 flat vectors are fits to the ratio of the extracted spectra to the superflat
@@
 \begin{figure}[tb]
+\figurenum{1}
+\epsscale{1.0}
+\plotone{plate_stats.ps}
+\figcaption{Observational summary of the 1656 publicly released
+SDSS spectroscopic
+plates observed between MJD 2451600 to 2453600.  Each plate is
+represented by 3 symbols showing the median RMS guiding residuals
+in the 11 guide fibers in arcseconds (diamonds), 
+the median seeing of the 11 guide star PSFs in arcseconds (stars), 
+and the effective (S/N)$^2$ of the plate observation
+normalized to a 1 hour exposure.  Vertical lines show the date of July 15 across
+the years of 2000 to 2005.
+\label{fig1}
+}
+\end{figure}
+
+\begin{figure}[tb]
+\figurenum{2}
+\epsscale{1.0}
+\plotone{spDiag2d-0412-52258.ps_1}
+\figcaption{A typical example of the common flat-field spectra, $SF$,
+for the b1 camera observations of plate 412.  The solid line is the best-fit
+B-spline function, and the red points are rejected 2-sigma outliers. 
+\label{fig2}
+}
+\end{figure}
+
+\begin{figure}[tb]
 \figurenum{7}

### 2006-04-01 doc/paper/paper2d.tex
Updated figures

@@
 
+\begin{figure}[tb]
+\figurenum{8}
+\epsscale{1.0}
+\plotone{spDiag2d-0412-52258_156.eps}
+\figcaption{An example of 1-d sky subtraction for camera r1 on SDSS
+plate 412.  The top panel shows data from 16 sky fibers corrected to airmass
+unity with vertical extent equal to 67\% errors.  The solid red line shows
+the best fit B-spline, with only wavelength as an independent variable.
+The bottom panel shows the residual of the fit in flux counts in the sky 
+spectra.
+\label{fig8}
+}
+\end{figure}
+
+\begin{figure}[tb]
+\figurenum{9}
+\epsscale{1.0}
+\plotone{spDiag2d-0412-52258.ps_153}
+\figcaption{An example of the wavelength corrections to match expected 
+sky emission line wavelength centers in a single r1 exposure.  The position
+residuals in spectral pixels are shown as measurements of 12 sky 
+emission lines for all 320 fibers.  The red positions show the initial 
+predicted positions based on the arc-lamp wavelength solutions, and the green
+positions are based on the updated solution after shifting and stretching to
+match the observed sky lines positions.  The residuals can be compared to a
+full pixel in the top left corner, and are much smaller than 5\%.
+\label{fig9}
+}
+\end{figure}
+
 \end{document}

### 2006-04-25 doc/paper/paper2d.tex
Added crude figure of spectrograph layout
small caption change
Update M67 plots.

@@
 specific idea of testing the SDSS velocity precision.
+This was plate 321 observed on MJD 51612.  
 130 stars were targetted from the radial velocity catalog
 of Mathieu \etal\ (1986), falling in the magnitude range of 8 to 14.
-The 19 stars brighter than $m=11$ were not plugged, since they
-would have saturated the detectors.  This was plate 321 MJD 51612.
+The 19 stars brighter than $m=11$ were intentionally not plugged, since they
+would have saturated the detectors, and another 8 stars appear not
+to have been plugged.  SDSS velocities for these stars were determined
+by fitting these spectra to all stars in the Elodie-2 stellar catalog,
+and choosing the best-fit Elodie star using $\chi^2$ minimization.  (ref???)
+The derived veocites have a median offset of $-5.5\kms$, with a sigma-clipped
+dispersion of $3.0 \kms$ (see Figure \ref{fig_m67}).
+There were three very discrepent velocities. (look these up???)
+% Fiber 224 at RA= 132.40396 Dec= 11.484270 v(CAT)= 13.30 v(Elodie)= 100.50532
+% Fiber 301 at RA= 132.01817 Dec= 12.167990 v(CAT)= 12.10 v(Elodie)= 104.35960
+
+Our expectation is that this plate represents a worst-case scenario
+for the zero-points in the wavelength scales, and therefore velocties.
+The exposure times were 30, 60 and 120 sec for this plate, as opposed
+to typical exposure times of 900 to 1500 sec on SDSS main survey plates.
+With the shorter exposures, there is not as much signal in the sky lines
+for computing the instrument flexure between the arc lamp spectra and
+the science exposures.
 
@@
 
-Below, I've listed the vacuum wavelength of some transitions:
+Below are listed the vacuum wavelength of some transitions:
 \begin{verbatim}
@@
 
-\begin{figure}[tb]
+\begin{figure}
 \figurenum{1}
@@
 
+\begin{figure}
+\figurenum{2}
+\epsscale{0.9}
+\plotone{layout.ps}
+\figcaption{Schematic layout of one of the SDSS spectrographs, from the 
+SDSS Project Book (http://www.astro.princeton.edu/PBOOK/spectro/spectro.htm).
+Four main components are labeled: A) the slithead, B) the collimator, 
+C) the beamsplitter, D) the grisms.
+\label{fig2}
+}
+\end{figure}
+
 \begin{figure}[tb]
-\figurenum{2}
+\figurenum{3}
 \epsscale{1.0}
@@
 B-spline function, and the red points are rejected 2-sigma outliers. 
-\label{fig2}
+\label{fig3}
 }
@@
 \begin{figure}[tb]
-\figurenum{7}
 \epsscale{0.9}
@@
 \figcaption{Measurements of spectral resolution in the four SDSS 
-spectrograph cameras.  For each camera, we should the median
+spectrograph cameras.  Shown for each camera is the median
 FWHM arc-line widths as a function of wavelength as 
@@
 the 10\% to 90\% ranges in each camera over all fibers and
-all plates through DR5.
-\label{fig7}
+all public data release plates through DR5.
+\label{fig_disp}
 }
@@
 
-\begin{figure}[tb]
+\begin{figure}
+\epsscale{0.9}
+\plotone{plot_m67.ps}
+\figcaption{
+SDSS velocities of 103 stars in or near the open cluster M67
+compared to published velocities in Mathieu \etal\  (1986).
+There is an offset of $5.0\kms$, and a sigma-clipped dispersion
+of $3.0\kms$.
+\label{fig_m67}
+}
+\end{figure}
+
+\begin{figure}
 \figurenum{8}
@@
 
-\begin{figure}[tb]
+\begin{figure}
 \figurenum{9}

### 2006-06-08 doc/paper/paper2d.tex
Add figures.

@@
 The typical flux distortion corrections amount to ???.
+
+\subsection{Tests of Spectrophotometry}
 
@@
 \epsscale{0.9}
-\plotone{layout.ps}
-\figcaption{Schematic layout of one of the SDSS spectrographs, from the 
-SDSS Project Book (http://www.astro.princeton.edu/PBOOK/spectro/spectro.htm).
-Four main components are labeled: A) the slithead, B) the collimator, 
-C) the beamsplitter, D) the grisms.
+\plotone{SpecSecFull.ps}
+\figcaption{Schematic layout of one of the SDSS double-spectrographs.
+The light is fed from 320 fibers entering at the slithead (A), which
+is in the center of the beam in this Schmidt camera design.
 \label{fig2}
@@
 
+\begin{figure}
+\figurenum{10}
+\epsscale{1.0}
+\plotone{plot_bestthru.ps}
+\figcaption{Throughput of the SDSS spectroscopic system,
+as measured relative to the photon density expected above
+the Earth's atmospheric by an aperture of 2.5-m with a 1.3-m
+obscuration by the secondary mirror baffling.  These curves
+are for plate 406 observed on MJD 51817.
+Solid curves are for spectrograph \#1, and dotted curves
+for spectrograph \#2.
+\label{fig10}
+}
+\end{figure}
+
 \end{document}

### 2006-06-09 doc/paper/paper2d.tex
Sky and LRG noise residual update

@@
 }
-\altaffiltext{1}{Massachussetts Institute of Technology,
-Boston, MA ????? \label{MIT}}
+\altaffiltext{1}{Department of Physics and Kavli Institute for Astrophysics 
+and Space Research, MIT, Cambridge, MA 02139 \label{MIT}}
 \altaffiltext{2}{Princeton University Observatory, Peyton Hall, Princeton,
@@
 width closely is closely represented by the reported variance array.
+
+\begin{table}
+\begin{tabular}{ccccccccccc}
+   -5   &  -4   &   -3  &   -2  &  -1   & 0 &  1   &  2 &  3   & 4 & 5 \\
+\hline
+ -15.19 & -6.65 & -3.13 & -1.86 & -0.89 &  0.047  & 0.98 &  1.95 &  3.22 &  6.74 & 15.29 \\
+ -15.76 & -7.31 & -3.35 & -1.95 & -0.93 &  0.041  & 1.01 &  2.04 &  3.44 &  7.40 & 15.85 \\
+ -16.37 & -8.52 & -3.72 & -2.04 & -0.98 & -0.014  & 0.95 &  2.01 &  3.69 &  8.50 & 16.35 \\
+ -16.77 &-10.43 & -4.43 & -2.21 & -1.05 & -0.063  & 0.93 &  2.08 &  4.30 & 10.31 & 16.65 \\
+\end{tabular}
+\end{table}
 
@@
 \epsscale{1.0}
-\plotone{spDiag2d-0412-52258_156.eps}
+\plotone{spDiag2d-0412-52258.ps_156}
 \figcaption{An example of 1-d sky subtraction for camera r1 on SDSS
@@
 \plotone{plot_bestthru.ps}
-\figcaption{Throughput of the SDSS spectroscopic system,
+\figcaption{ Throughput of the SDSS spectroscopic system,
 as measured relative to the photon density expected above
@@
 
+\begin{figure}
+\figurenum{11}
+\epsscale{1.0}
+\plottwo{plot_sky_reshist.ps}{plot_lrg_reshist.ps}
+\figcaption{ Distribution of SDSS flux residuals normalized 
+by the reported pipeline inverse variances.  The left plot shows
+the distribution of 214 million sky pixels in reductions from both
+Version 4 (upper), and Version 5 (lower) of spectro-2d.  
+The best fit functional form is shown as a solid red line (see text).
+The filled region shows the area between the observed version 5 distribution 
+of sky flux residuals and a gaussian with identical 1-sigma confidence limits.
+The right plot is the same but shows the distribution of approximately 
+550 million pixels drawn from residuals with respect to best fit models of
+luminous red galaxies.
+\label{fig11}
+}
+\end{figure}
+
 \end{document}

### 2006-06-09 doc/paper/paper2d.tex
Update.
Fixed some figures

@@
 in Gunn et al. 2006.  The switch from the imaging camera to a spectroscopic
-cartridge takes an average of xx minutes, while a switch from one cartridge
-to another takes merely x minutes.  Once the cartridge is mounted, the telescope is slewed to the coordinates of the plate center.  
+cartridge takes an average of 30 minutes, while a switch from one cartridge
+to another takes 10 minutes.  Once the cartridge is mounted, 
+the telescope is slewed to the coordinates of the plate center.  
 Before acquisition of the guide stars, the 8 flat-field petals are closed and
@@
 being readout, the flat-field petals are opened, and guide stars are acquired
-through the 11 coherent guide bundles (9 are this big and 2 are even bigger??).
+through the 11 coherent guide bundles (9 are 7$"$ in diameter and 
+2 are even larger at 11$"$).
 With at least 4 guide stars acquired, closed loop guiding is started, and
-usually converges to stable guiding within 2 minutes??
+usually converges to stable guiding within 2 minutes.
 The plate scale of the SDSS telescope can be adjusted, and the guiding software
@@
 if all 4 cameras have a cumulative (S/N)$^2$ greater than 15 at 
-fidicual magnitudes of ?? and ??.
+fidicual magnitudes (what the heck is a fiducial magnitude??)
+of 20.33 and 20.06 in SDSS g and i filters respectively.
 
@@
 sorted in order of fiber number with background parameters included at the
-end of the list.  This facilitates and banded matrix inversion, and 
+end of the list.  This facilitates a banded matrix inversion, and 
 significantly increases the speed of Cholesky decomposition step in each
@@
 minimum number of free parameters.  In the end, we found the best 
-results using a simple Gaussian or a slightly steeper profile 
-characterized by $e^{-|x|^3/\sigma^3}$.  The profiles are defined to be 
+results using a simple Gaussian (noted as PROFTYPE 1 in the primary HDU header) 
+or a slightly steeper profile characterized by 
+the average of a profile with $e^{-|x|^3/\sigma^3}$ and a Gaussian
+with the same sigma (PROFTYPE 3).  The profiles are defined to be 
 normalized to unity, so the amplitude of the profile fit is a 
@@
 Each flat-field frame is extracted twice, once assuming a Gaussian
-profile and again assuming the steeper $|x|^3$ profile.  In both cases,
+profile and again assuming the steeper PROFTYPE 3 profile.  In both cases,
 the width is allowed to vary freely.  The total number of free parameters
@@
 fit in wavelength bins separated by breakpoints determined above.  For each bin
-where inverse the 67th percentile of $\chi^2$ is greater than 1, we rescale 
+where the 67th percentile of $\chi^2$ is greater than 1, we rescale 
 all the inverse variance values down by the 67th percentile value of $\chi^2$.
@@
 
-The accuracy of the SDSS spectral sky subtraction has been tested in recent
-scientific analyses using large numbers of background-limited spectra.   
-The distribution of flux residuals between spectral observations 
-on separate nights has been shown to be 8\% larger in width than the average expected
-from the reported errors (McDonald et al. 2004, Burgess \& Burles 2006).
-The spectra of faint, background-limited galaxies observed with SDSS exhibit a 
-distribution of residuals about the best fit spectral models described by the 
-sum of a
-Gaussian and Laplace distribution (Bolton et al. 2004).  
-The influence of the exponential tail is greatly enhanced in the vicinity of strong sky 
-emission lines representative of systematic errors in the brightest lines, but the gaussian 
-width closely is closely represented by the reported variance array.
+The accuracy of the SDSS spectral sky subtraction can been tested by examining
+the residuals in final reduced spectra, where spectra of blank sky, high-z QSOs
+and luminous red galaxies have been investigated to date.
+For early releases of spectroscopic data, reduced by version 4 and include
+data releases 4 and earlier, there have been mulitple confirmations of an
+underestimation of the true variance as reported in the output of the 
+spectroscopic pipeline.  By comparing multiple observations of the same
+high-z QSOs in studies of the Lyman-$\alpha$ forest, the average variance was
+shown to be underestimated by approximately 16\% (McDonald et al. 2004).  
+A fraction of the effect arises due to large tails in the distribution 
+of the noise and the remainder from the predicted FWHM of the noise 
+distribution.  Switzer (2004) identified two causes for the underestimates,
+citing both the awkward treatment of large sky residuals relative to the
+supersky and the corresponding correction described above, and also a general
+overestimation of the gains applied for the 8 CCD amplifiers.
+
+Using the sample of plates here, we investigate the distribution of the 
+noise in a large sample of sky fibers as
+well as a sample of model fit luminous red galaxies.  Inclusion of
+the latter sample is motivated by earlier studies which showed 
+the spectroscopic residuals about the best fit galaxy models 
+can be described by a sum of Gaussian and Laplace 
+distributions (Bolton et al. 2004).  We construct histograms of the flux
+pixels directly from the SDSS spectroscopic products for two versions
+of the software (versions 4 \& 5).  In the first sample, flux measurements
+from the sky fibers are shown relative to zero and normalized by their reported
+inverse variances 
+(${\rm residual} = {\rm flux} \, \sqrt{\rm inverse variance}$).  
+In the second sample, residuals in galaxy spectra, 
+which were targetted as luminous red galaxies, were constructed by subtracting
+the best fit galaxy model and then normalizing by the inverse variance.
+
+The results are shown in Figure 11, and also summarized in Table 1.
+In Fig. 11, we present the distributions (histograms) as 
+probability densities (normalized to unit probability) on a logarithmic 
+scale for spectro\_2d versions 4 \& 5.
 
@@
 
+\bibitem[Schoenberg 1946]{} Schoenberg, I. J. 1946, Quart. Appl. Math., 4, 45
+
 \bibitem[Schlegel, Finkbeiner \& Davis 1998 (SFD)]{sfd98} 
@@
 In many of our algorithms we utilize a class of fitting functions known
-as B-splines (de Boor 1978).  These are very flexible and efficient 
+as B-splines ("basis splines", de Boor 1978, Schoenberg 1946).  
+These are very flexible and efficient 
 continous functions which have been applied to numerous spectroscopic
@@
 \figurenum{3}
-\epsscale{1.0}
-\plotone{spDiag2d-0412-52258.ps_1}
+\centering
+\leavevmode
+\includegraphics[width=\columnwidth,clip]{spDiag2d-0412-52258.ps_1}
+%\epsscale{1.0}
+%\plotone{spDiag2d-0412-52258.ps_1}
 \figcaption{A typical example of the common flat-field spectra, $SF$,
@@
 The range of resolution is shown as error bars representing
-the 10\% to 90\% ranges in each camera over all fibers and
+the 10\% to 90\% range in each camera over all fibers and
 all public data release plates through DR5.
@@
 \epsscale{1.0}
-\plotone{spDiag2d-0412-52258.ps_156}
+\plotone{skysub_1d.ps}
 \figcaption{An example of 1-d sky subtraction for camera r1 on SDSS
@@
 \figurenum{9}
-\epsscale{1.0}
-\plotone{spDiag2d-0412-52258.ps_153}
+\centering
+\leavevmode
+\includegraphics[width=\columnwidth,clip]{spDiag2d-0412-52258.ps_153}
+%\plotone{spDiag2d-0412-52258.ps_153}
 \figcaption{An example of the wavelength corrections to match expected 
@@
 match the observed sky lines positions.  The residuals can be compared to a 
-full pixel in the top left corner, and are much smaller than 5\%.
+full pixel in the top left corner, and are much smaller than 5\% of a pixel
+width.
 \label{fig9}
@@
 obscuration by the secondary mirror baffling.  These curves
-are for plate 406 observed on MJD 51817.
+are for plate 406 observed on MJD 51817, which was observed
+in good-seeing photometric conditions shortly after
+re-aluminization of the mirrors.
 Solid curves are for spectrograph \#1, and dotted curves
@@
 by the reported pipeline inverse variances.  The left plot shows
-the distribution of 214 million sky pixels in reductions from both
-Version 4 (upper), and Version 5 (lower) of spectro-2d.  
+the distribution as a histrogram of 214 million sky pixels in 
+reductions from both version 4 (DR4 and earlier, upper) 
+and version 5 (lower) of spectro\_2d.  
 The best fit functional form is shown as a solid red line (see text).

### 2006-06-10 doc/paper/paper2d.tex
Update throughput vs. time.
Updated text with Jill's comments up to flux calibration

@@
 The Sloan Digital Sky Survey (SDSS) fiber-fed spectrographs have been used
-to obtain spectra for $600,000$ galaxies, QSOs and stars.
+to obtain spectra for ~$600,000$ galaxies, quasars and stars
+with a resolution of ~2000 over the wavelength range 3800\AA -9600\AA.
 We present the algorithms for reducing the raw data to
@@
 The wavelength calibration is shown to have an accuracy of $3 \kms$,
-and the broad-band spectrophotometry has an accuracy of better than $6\%$.
+and the broad-band spectrophotometry to have an accuracy of better than $6\%$.
 
@@
 redshift survey of up to a million galaxies and 100,000 quasars
-selected from the imaging survey \cite{york00}.  The primary 
+selected from the imaging survey (York et al 2000).  The primary 
 science goals of the project are to provide the data to investigate the
@@
 to June 15, 2005 (MJD 2453536).  Over this timespan, which includes public data 
-releases EDR through DR5, SDSS obtained survey quality spectroscopic data on
+releases EDR through DR5 (Stoughton et al ??, Abazajian et al ??, 
+Adelman-McCarthy et al)
+, SDSS obtained survey quality spectroscopic data on
 1656 fiber pluggings (640 fibers each in 1575 unique plates).   
@@
 Mean exposure time per plate is 3914 seconds, and the median is 3302 seconds.
-In Figure 1 we show three of the observing diagnostics, on the top is
-the effective signal-to-noise of the whole plate normalized to 3600 second
-exposure time, the median seeing in arcseconds and the rms guide probe offset
-in arcseconds are shown below.  The abscissa is in days after Feb 26, 2000
-(MJD 2451600).
+In Figure 1 we show three of the data quality diagnostics. This figure shows
+the effective signal-to-noise ratio of each spectroscopic plug plate 
+normalized to 3600 seconds exposure time, the median seeing in arcseconds,
+and the RMS guide probe offset
+in arcseconds, versus time in days after Feb 26, 2000 (MJD 2451600).
 
 The data reduction and products described herein are very similar to
-those produced by the SDSS, with the most significant differences
-in our treatment of spectro-photometric calibrations.
-The data are released
+those produced by the official SDSS data releases, 
+with the most significant differences begin in our treatment of 
+spectro-photometric calibrations.  The data as processed as described
+are released
 with this paper and may be accessed via the WWW.\footnote{Full details
@@
 The SDSS spectroscopic survey was carried out with twin multi-fiber 
-spectrographs mounted at the back of the SDSS telescope (see Gunn et al.
-2006).  One of nine spectrograph cartidges can be mounted for a single
+double spectrographs mounted at the back of the SDSS telescope 
+(see Gunn et al.  2006).  
+One of nine spectrograph cartidges can be mounted for a single
 spectroscopic observing sequence.  The cartridge holds a single aluminum
@@
 
-The fibers are 180 $\mu m$ in diameter and subtend 2.97" at Cassegrain focus.
+The fibers are 180 $\mu m$ in diameter and subtend $2.97"$ at Cassegrain focus.
 The fibers accept the f/5 beam directly and, due to focal degradation, input
@@
 The output light from the fiber is collimated with a spherical mirror
-(radius of curvature is 1264 mm), which is aligned by 3 motor mounts to
-control focus, tip \& tilt.  A left/right hartmann mask 
+(of radius of curvature 1264 mm), which is aligned by 3 motor mounts to
+control focus, tip and tilt.  A left/right Hartmann mask 
 covers the collimator to act as both a spectrograph shutter and traditional 
-shack-hartmann mask for focus tests.  The collimator in spectrograph 2
+Shack-Hartmann mask for focus tests.  The collimator in spectrograph 2
 was coated with an enhanced silver coating and delivers slightly higher
@@
 The collimated beam encounters a 40mm thick dichroic (beamsplitter) which
-reflect light blueward of 6000$\AA$.  The blue reflected light is dispersed
-with a 640 l/mm grism with zero deviation at 4960$\AA$.  The red transmitted
-light is dispersed with a 440 l/mm grism with zero deviation at 7400$\AA$.
+reflect light blueward of 6000\AA.  The blue reflected light is dispersed
+with a 640 l/mm grism with zero deviation at 4960\AA.  The red transmitted
+light is dispersed with a 440 l/mm grism with zero deviation at 7400\AA.
 The dispersed light in each arm of the spectrograph enters a transmission camera
-wth a 240mm focal length (demagnification of 2.5) which is tuned for the 
-corresponding range of wavelengths.  Each of the 4 camera produce RMS 
+with a 240mm focal length (demagnification of 2.5) which is tuned for the 
+corresponding range of wavelengths.  Each of the 4 cameras produce RMS 
 spot sizes of approximately 20$\mu m$ over a 16.5$\degree$ diameter 
-f.o.v. delivered to a single thinned Tektronix 2048x2048x24$\mu m$ CCD.
+field-of-view delivered to a single thinned Tektronix 2048x2048x24$\mu m$ CCD
+(see Gunn et al. 2006).
 An external focus ring is adjustable by hand for each CCD camera to accomodate
 differential focus changes not corrected with the common collimator.
-The four spectrograph cameras are labeled "b1, r1, b2, r2".
+The four spectrograph cameras are labeled $``$b1, r1, b2, r2", where
+1 and 2 refer to the spectrographs and b and r to their blue and red sides.
 
@@
 in cartridges, plugged with 640 science fibers and 11 guide fibers. 
-The position of the 640 fibers along each of the two fiber slitheads
-are mapped by the 8th wonder of the world,
-the SDSS plugplate fiber mapper (Schlegel and Finkbeiner).
+The positions of the 640 fibers along each of the two fiber slitheads
+%are mapped by the 8th wonder of the world,
+by 
+the SDSS plugplate fiber mapper (Schlegel and Finkbeiner, unpublished).
 All usable spectroscopic data obtained for a single plugging, 
@@
 output spectrum is uniquely specified by the plate number, the first night
-of observations (in MJD), and the science fiber number.
+of observations\footnote{On occasion, to accumulate enough signal, a given
+plate/plugging configuration may be exposed on more than one night.} (in MJD), 
+and the science fiber number.
 
 A plate is mounted in the telescope through a efficient procedure described
-in Gunn et al. 2006.  The switch from the imaging camera to a spectroscopic
+by Gunn et al (2006).  The switch from the imaging camera to a spectroscopic
 cartridge takes an average of 30 minutes, while a switch from one cartridge
@@
 Neon-Mercury-Cadmium arc exposure are obtained.  While the arc-lamp frame is
-being readout, the flat-field petals are opened, and guide stars are acquired
+being read out, the flat-field petals are opened, and guide stars are acquired
 through the 11 coherent guide bundles (9 are 7$"$ in diameter and 
@@
 The plate scale of the SDSS telescope can be adjusted, and the guiding software
-suggest a scale change if one is required.  Once close loop 
+suggests a scale change if one is required.  Once close loop 
 guiding has settled, the first science exposure is started.  The typical 
@@
 not unexpected based on conditions and time constraints.  A second exposure
-is started immediately after the first science exposure is complete and readout.A fast linux workstation onsite copies the data from the data acquisition host,
+is started immediately after the first science exposure is complete and read
+out.
+A fast Linux workstation onsite copies the data from the data acquisition host,
 and automatically performs a quick spectral extraction with a subset of the 
 steps explained here.  The results are posted to sos.apo.nmsu.edu, and
-the SDSS observers can check the S/N obtained in each of the 4 spectroscopic
+the SDSS observers can check the signal-to-noise ratio (S/N) 
+obtained in each of the 4 spectroscopic
 cameras in real-time.  A plate is considered to have met SDSS survey standards
@@
 We utilize an optimal extraction procedure to transform a 2-dimensional image of
-the fiber spectrograms to a set of 320 1-dimensional spectra (per spectrograph).
-We use the same procedure on all processed frames, 
+the fiber spectrograms to a set of 320 one-dimensional spectra 
+(per spectrograph).  We use the same procedure on all processed frames, 
 and will describe the details in processing the flat-field frames, the
@@
 end of the list.  This facilitates a banded matrix inversion, and 
-significantly increases the speed of Cholesky decomposition step in each
+significantly increases the speed of the Cholesky decomposition step in each
 iteration of the extraction.
@@
 
-\item{A valid estimate of the error in flux measurement at each spectral pixel.}
+\item{A valid estimate of the error in a flux measurement at 
+each spectral pixel can be made.}
 
@@
 A flat-field frame is rejected by the pipeline if one of the
-following conditions is met:  1) more the 2\% of the CCD pixels are labeled
-bad by sdssproc, 2) more than 100 rows of the image have more than 20 saturated
-pixels, 
+following conditions is met:  1) more than 2\% of the CCD pixels are labeled
+bad by {\tt sdssproc} (see below), 
+2) more than 100 rows of the image have more than 20 saturated pixels, 
 3) the 80th percentile count level is less than 1000 counts, 
@@
 in CCD pixel coordinates.
-This is performed by identifying peaks in a cross-sectional cut of across 
+This is performed by identifying peaks in a cross-sectional cut across 
 the center row (actually a median cross-section of the central 9 rows).  
@@
 and implemented for the extraction of the remaining frames.  Also retained
-is a smooth polynomial fit the average profile width in each fiber bundle
+is a smooth polynomial fit to the average profile width in each fiber bundle
 as a function of CCD row number.
@@
 A representative spectrum is calculated as the median extracted spectrum
-of the 5 closest "unflagged" spectra falling near the central columns 
+of the 5 closest $``$unflagged" spectra falling near the central columns 
 of the arc-lamp image.  This median arc-spectrum is matched with 
@@
 is declared unacceptable ("BAD") if any of these following criteria are met:
-5) all 8 arc-lamps must be on, 4 Neon lamps and 4 Mercury-Cadmium lamps,
+5) not all 8 arc-lamps are on, 4 Neon lamps and 4 Mercury-Cadmium lamps,
 6) the initial cross-correlation coefficient is less than 0.5,
@@
 \end{equation} 
-
 where $A$ is a overall normalization in counts, 
@@
 
-The common normalized response, $SF$, is specified as a B-spline
-function with breakpoints spaced approximately every native pixels.
+The common normalized response, $SF$, is specified as a B-spline (basis spline)
+function with breakpoints spaced at approximately the native pixel spacing.
 An example of normalized individual fiber spectra 
-and the resulting B-spline fit are shown in Figure 2.  The individual fiber
+and the resulting B-spline fit is shown in Figure 3.  The individual fiber
 flat vectors are fits to the ratio of the extracted spectra to the superflat
-(define above).  The individual vector fits are also B-splines with breakpoints
+(defined above).  The individual vector fits are also B-splines with breakpoints
 spaced approximately every 5 native pixels.  The final set of flat vectors are
-stored as an image with dimensions Nfiber x Npixels.
+stored as an image with dimensions of Nfiber,Npixels.
 
@@
 We reject a science exposure outright if 1) more than 10\% of the pixels
-are flagged as bad, 2) more than 100 rows are have more than 20 saturated
+are flagged as bad, 2) more than 100 rows have more than 20 saturated
 pixels, 3) the 25th
@@
 are passed directly to a test for overly bright fibers. Fiber spectra are 
-identified as bright (a.k.a. "WHOPPING") if the median counts per pixel in the 
+identified as bright (a.k.a. $``$WHOPPING") 
+if the median counts per pixel in the 
 boxcar extraction are greater than 10000 counts above the median in the 
 frame.  In addition, only the brightest fiber in a spatially contiguous set of 
-bright fibers is labelled "WHOPPING".  Each whopping fiber adds an additional
+bright fibers is labelled $``$WHOPPING".  Each whopping fiber adds an additional
 background parameter to the fit per row, which is an exponential profile
@@
 A pixelmask vector is created for each fiber spectrum to be extracted.
-See Appendix ?? for a description of the mask bits and their use.
+See Appendix B for a description of the mask bits and their use.
 
@@
 to match the identified sky line positions.  
-Examples of the shift corrections are shown in Figure ??.  
+Examples of the shift corrections are shown in Figure 9.  
 The widths of the identified sky lines are also measured with the same method
@@
 The wavelength solution is corrected to both vacuum wavelengths and the
-barycenter velocity frame.
+solar system barycenter velocity frame.
 
@@
 and re-scaled errors is called 3 times successively.  After the 1st call,
-any sky fibers which have a median $\chi^2$ per pixel greater than 2 is 
-masked as a 'BADSKYFIBER' and the corresponding bit is set in the fibermask
+any sky fiber which has a median $\chi^2$ per pixel greater than 2 is 
+masked as a $``$BADSKYFIBER" and the corresponding bit is set in the fibermask
 array.  If any fibers are masked, the sky subtraction is iterated a 2nd time.
@@
 \begin{table}
-\begin{tabular}{ccccccccccc}
-   -5   &  -4   &   -3  &   -2  &  -1   & 0 &  1   &  2 &  3   & 4 & 5 \\
+\begin{tabular}{cccccccccccc}
 \hline
- -15.19 & -6.65 & -3.13 & -1.86 & -0.89 &  0.047  & 0.98 &  1.95 &  3.22 &  6.74 & 15.29 \\
- -15.76 & -7.31 & -3.35 & -1.95 & -0.93 &  0.041  & 1.01 &  2.04 &  3.44 &  7.40 & 15.85 \\
- -16.37 & -8.52 & -3.72 & -2.04 & -0.98 & -0.014  & 0.95 &  2.01 &  3.69 &  8.50 & 16.35 \\
- -16.77 &-10.43 & -4.43 & -2.21 & -1.05 & -0.063  & 0.93 &  2.08 &  4.30 & 10.31 & 16.65 \\
+sigma &   -5   &  -4   &   -3  &   -2  &  -1   & 0 &  1   &  2 &  3   & 4 & 5 \\
+\hline
+\hline
+Sky V5 & -15.19 & -6.65 & -3.13 & -1.86 & -0.89 &  0.047  & 0.98 &  1.95 &  3.22 &  6.74 & 15.29 \\
+Sky V4 & -15.76 & -7.31 & -3.35 & -1.95 & -0.93 &  0.041  & 1.01 &  2.04 &  3.44 &  7.40 & 15.85 \\
+LRG V5 & -16.37 & -8.52 & -3.72 & -2.04 & -0.98 & -0.014  & 0.95 &  2.01 &  3.69 &  8.50 & 16.35 \\
+LRG V4 & -16.77 &-10.43 & -4.43 & -2.21 & -1.05 & -0.063  & 0.93 &  2.08 &  4.30 & 10.31 & 16.65 \\
+\hline
 \end{tabular}
@@
 with at least $90\%$ of the pixels good in the spectra.
-For stars, we use the PSF fluxes as measured by PHOTO.
+For stars, we use the PSF fluxes as measured by
+the SDSS imaging as reduced by the photometric pipeline PHOTO 
+(Lupton et al ??).
 For galaxies, we rescale the fiber fluxes as measured by PHOTO,
 rescaled by one multiplicative factor in each band such that the
-PSF and fiber fluxes agree for stars.  Typically, that mupliplies
+PSF and fiber fluxes agree for stars.  Typically, that multiplies
 the PHOTO fiber fluxes by factors of 1.37 in any band.
@@
 %------------------------------------------------------------------------------
-%\section{SUMMARY}
+\section{SUMMARY}
+
+Other pipelines using the data.
+
+Known problems with the data.
+
+Still need to finish testing velocities and sdssproc description.
 
@@
 \bigskip
-We are very grateful to Michael Strauss for
-useful and helpful comments on the manuscript.
+We are very grateful to Jill Knapp, Craig Loomis, and Michael Strauss for
+useful and helpful comments on the manuscript.  Adam Bolton 
+graciously supplied some unpublished results on the measured noise residuals.
+
 Funding for the creation and distribution of the SDSS Archive has been
@@
 \figurenum{10}
-\epsscale{1.0}
-\plotone{plot_bestthru.ps}
+\centering
+\leavevmode
+\includegraphics[width=0.8\columnwidth,angle=90]{plot_bestthru.ps}
+%\epsscale{1.0}
+%\plotone{plot_bestthru.ps}
 \figcaption{ Throughput of the SDSS spectroscopic system,
@@
 
+\begin{figure}
+\figurenum{10}
+\epsscale{1.0}
+\plotone{plot_thru.ps}
+\figcaption{ History of throughput for the red channel on spectrograph \#2
+convolved with the SDSS i-band filter.
+The median throughput is 12\%.
+Vertical lines denote four times when the primary mirror was realuminized.
+\label{fig_thru_history}
+}
+\end{figure}
+
+\begin{figure}
+\figurenum{11}
+\epsscale{1.0}
 \begin{figure}
@@
 
+\begin{figure}
+\figurenum{12}
+\epsscale{1.0}
+\plotone{sky_cross.ps}
+\figcaption{The fractional contribution of the flux from 
+spectroscopic fibers neighboring sky fibers.  We fit for the unweighted 
+amplitude using the extracted neighbors as a template.  We include 93837
+spectroscopic neighbors of sky fibers with $r$ fiber magnitude between 14.5
+and 20.0.  The median contamination of the sky fiber is 0.0013 of the 
+flux level of the neighbor.  The histrograms show the 25\%, 50\% and 75\%
+quartiles as a function of magnitude.
+\label{fig12}
+}
+\end{figure}
+
 \end{document}

### 2006-11-09 doc/paper/paper2d.tex
Update.
Still working, with references

@@
 \emph{Subject headings: }
-???
+instrumentation: spectrographs --- techniques: spectroscopic --- 
+methods: data analysis --- surveys -- galaxies: distances and redshifts
 \end{abstract}
@@
 In this paper, we describe the details of the SDSS 2-d spectroscopic pipeline
-(idlspec2d) including steps starting with processing of the raw CCD frames to final
-flux calibration and spectral co-addition.  Downstream pipelines which
-calculate redshifts, spectral types, emission and absorption line identification,
-and velocity dispersions are described elsewhere.
+(idlspec2d) including steps starting with processing of the raw CCD frames 
+to final flux calibration and spectral co-addition.  Downstream pipelines 
+which calculate redshifts, spectral types, emission and absorption line 
+identification, and velocity dispersions are described elsewhere.
 As of writing the 2-d pipeline has been run successfully on all spectroscopic
-data obtained in the course of the survey from February 28, 2000 (MJD 2451602) 
+data obtained in the course of the survey from February 28, 2000 
+(Modified Julian Data, MJD 2451602) 
 to June 15, 2005 (MJD 2453536).  Over this timespan, which includes public data 
-releases EDR through DR5 (Stoughton et al ??, Abazajian et al ??, 
-Adelman-McCarthy et al)
+releases EDR through DR5 (Stoughton et al 2002, 
+Abazajian et al 2003, 2004, 2005, Adelman-McCarthy et al 2006)
 , SDSS obtained survey quality spectroscopic data on
@@
 Mean exposure time per plate is 3914 seconds, and the median is 3302 seconds.
+;
+;   Should we include a histogram of exposure times??
+;
+
 In Figure 1 we show three of the data quality diagnostics. This figure shows
@@
 
-In Figure 1, we show a top-view schematic of one the spectrographs, and
-will detail the light path below starting at the slithead.
-The output light from the fiber is collimated with a spherical mirror
+In Figure 1, we show a top-view schematic of one the spectrographs.  Here
+we provide a short summary of the light path starting at the slithead,
+further details can be found in Uomoto et al. (20??) and Gunn et al. (2006).
+The output light from the fiber is collimated by a spherical mirror
 (of radius of curvature 1264 mm), which is aligned by 3 motor mounts to
@@
 The collimated beam encounters a 40mm thick dichroic (beamsplitter) which
-reflect light blueward of 6000\AA.  The blue reflected light is dispersed
+reflects light blueward of 6000\AA.  The blue reflected light is dispersed
 with a 640 l/mm grism with zero deviation at 4960\AA.  The red transmitted
@@
 cameras in real-time.  A plate is considered to have met SDSS survey standards
-if all 4 cameras have a cumulative (S/N)$^2$ greater than 15 at 
-fidicual magnitudes (what the heck is a fiducial magnitude??)
-of 20.33 and 20.06 in SDSS g and i filters respectively.
+if all 4 cameras have a cumulative (S/N)$^2 > 15$ at 
+fiber magnitudes of 20.33 and 20.06 in SDSS g and i filters respectively.
+
+\section{IMAGE PROCESSING}
+\label{sec_process}
+
+sdssproc:
+
+1000 lines of IDL code...
+
+read opEcalib
+     op something else
+
+Read file header, require...
+mention spHdrFix here.
+
+\section{SPATIAL TRACING}
+\label{sec_tracing}
+
+trace320, Trace crude, and trace fitting
 
@@
 in tests of the flat-field images.  In addition, an arc-lamp exposure
-is declared unacceptable ("BAD") if any of these following criteria are met:
-5) not all 8 arc-lamps are on, 4 Neon lamps and 4 Mercury-Cadmium lamps,
+is declared unacceptable (``BAD") if any of these following criteria are met:
+5) not all 8 arc-lamps are on: 4 Neon lamps and 4 Mercury-Cadmium lamps,
 6) the initial cross-correlation coefficient is less than 0.5,
@@
 compared to the observed sky emission line shapes show less than a 1\%
-difference on average (McDonald et al. 2004).
+difference on average (McDonald et al. 2006).
 
@@
 high-z QSOs in studies of the Lyman-$\alpha$ forest, the average variance was
-shown to be underestimated by approximately 16\% (McDonald et al. 2004).  
+shown to be underestimated by approximately 16\% (McDonald et al. 2006).  
 A fraction of the effect arises due to large tails in the distribution 
@@
 inverse variances 
-(${\rm residual} = {\rm flux} \, \sqrt{\rm inverse variance}$).  
+(${\rm residual} = {\rm flux} \, \sqrt{\rm 1}$).  
 In the second sample, residuals in galaxy spectra, 
@@
 $$ r-i = 0.10 $$
-This is meant to select those stars closest in color to BD +17 4708,
-which is computed to have a similar color (0.93, 0.28, 0.01, courtesy
-of David Hogg, priv. comm.).
+in order to select those stars closest in color to BD +17 4708 
+(0.93, 0.28, 0.01, courtesy of David Hogg, priv. comm.).
 From these stars, we select 8 on each plate in the magnitude
@@
 \begin{thebibliography}{DUM}
+
+\bibitem[McDonald et al.(2006)]{2006ApJS..163...80M} McDonald, P., et al.\ 
+2006, \apjs, 163, 80
+
+\bibitem[Stoughton et al.(2002)]{2002AJ....123..485S} Stoughton, C., et 
+al.\ 2002, \aj, 123, 485
+
+\bibitem[Adelman-McCarthy et al.(2006)]{2006ApJS..162...38A} 
+Adelman-McCarthy, J.~K., et al.\ 2006, \apjs, 162, 38
+
+\bibitem[Abazajian et al.(2005)]{2005AJ....129.1755A} Abazajian, K., et 
+al.\ 2005, \aj, 129, 1755
+
+\bibitem[Abazajian et al.(2004)]{2004AJ....128..502A} Abazajian, K., et 
+al.\ 2004, \aj, 128, 502
+
+\bibitem[Abazajian et al.(2003)]{2003AJ....126.2081A} Abazajian, K., et 
+al.\ 2003, \aj, 126, 2081
 
@@
 \epsscale{1.0}
-\begin{figure}
-\figurenum{11}
-\epsscale{1.0}
 \plottwo{plot_sky_reshist.ps}{plot_lrg_reshist.ps}

### 2011-03-14 doc/paper/paper2d.tex
Testing Access
Updates to the image processing and spatial tracing. The current trunk code does not represent what was used in legacy reductions, need to find the appropriate revision to finish the spatial tracing section.

@@
 
-\title{The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline}
+\title{The Sloan Digital Sky Survey: Two-dimensional Spectroscopic Reduction Pipeline}
 
 \author{
-S. Burles\altaffilmark{\ref{MIT}}
-and D. J. Schlegel\altaffilmark{\ref{Princeton}}
+Scott Burles\altaffilmark{\ref{email}}
+and David J. Schlegel\altaffilmark{\ref{Princeton}}
 }
-\altaffiltext{1}{Department of Physics and Kavli Institute for Astrophysics 
-and Space Research, MIT, Cambridge, MA 02139 \label{MIT}}
+\altaffiltext{1}{sburles@gmail.com \label{sburles}}
 \altaffiltext{2}{Princeton University Observatory, Peyton Hall, Princeton,
@@
 with a resolution of ~2000 over the wavelength range 3800\AA -9600\AA.
-We present the algorithms for reducing the raw data to
+We present the pipeline design and algorithms for reducing the raw data to
 wavelength-calibrated, flux-calibrated one-dimensional spectra.
+It is designed to automatically reduce survey quality data sets and output 
+uniformly processed results.  No human intervention is required in any step 
+of the pipeline beyond invoking the top level command pointing to the 
+applicable data.
 The wavelength calibration is shown to have an accuracy of $3 \kms$,
 and the broad-band spectrophotometry to have an accuracy of better than $6\%$.
+What about reported errors and their accuracy/robustness?
+The reported errors capture the statistical and the expected have
 
@@
 science goals of the project are to provide the data to investigate the
-large scale structure of the Universe and other extragalactic science.
+large scale structure of the Universe as well as other extragalactic and 
+galactic science.
 
@@
 (idlspec2d) including steps starting with processing of the raw CCD frames 
-to final flux calibration and spectral co-addition.  Downstream pipelines 
-which calculate redshifts, spectral types, emission and absorption line 
-identification, and velocity dispersions are described elsewhere.
+to final flux calibration and spectral co-addition\footnote{By SDSS, 
+we specifically refer to the reduction pipeline for SDSS-I and the 
+legacy reductions for SDSS-II}.  The goal of the two-dimensional pipeline is to
+produce the best empirical estimate of spectral intensity as a function of 
+wavelength for each object observed by the SDSS spectroscopic system.  
+Downstream pipelines which calculate redshifts, spectral types, 
+emission and absorption line identification, and velocity dispersions 
+are described elsewhere (citations??).
+?? Update the numbers below to include data through DR7 or DR8 ??
 As of writing the 2-d pipeline has been run successfully on all spectroscopic
 data obtained in the course of the survey from February 28, 2000 
-(Modified Julian Data, MJD 2451602) 
-to June 15, 2005 (MJD 2453536).  Over this timespan, which includes public data 
+(Modified Julian Data, MJD 51602) to June 15, 2005 (MJD 53536).  
+Over this timespan, which includes public data 
 releases EDR through DR5 (Stoughton et al 2002, 
@@
 ;   Should we include a histogram of exposure times??
+;   No, probably not.
 ;
@@
 and the RMS guide probe offset
-in arcseconds, versus time in days after Feb 26, 2000 (MJD 2451600).
+in arcseconds, versus time in days after Feb 26, 2000 (MJD 51600).
 
@@
 are released
-with this paper and may be accessed via the WWW.\footnote{Full details
-are at \texttt{http://spectro.princeton.edu}}
+with this paper and may be accessed online.\footnote{Full details
+are at (should we point to DR sdss.org link??) \texttt{http://spectro.princeton.edu}}
 Future data releases will also be accessible at this site.
+
+In Section ??, we do this, in the next section we do this, and in the final sections we do this and that??
 
@@
 an f/4 exit beam into each of the spectrographs.  The fibers are mounted in
-bundles of 20 with inter-fiber spacings of 390 $\mu m$.  Sixteen bundles are
+linear bundles of 20 with inter-fiber spacings of 390 $\mu m$.  
+Sixteen bundles are
 mounted on a single slithead with a radius of curvature of 640 mm and have
@@
 by 
-the SDSS plugplate fiber mapper (Schlegel and Finkbeiner, unpublished).
+the SDSS plugplate fiber mapper (Schlegel and Finkbeiner, unpublished, ??should we use another reference??).
 All usable spectroscopic data obtained for a single plugging, 
@@
 output spectrum is uniquely specified by the plate number, the first night
-of observations\footnote{On occasion, to accumulate enough signal, a given
-plate/plugging configuration may be exposed on more than one night.} (in MJD), 
-and the science fiber number.
+of observations (in MJD)\footnote{On occasion, to accumulate enough signal, a given
+plate/plugging configuration may be exposed on more than one night.} 
+, and the science fiber number (001-640).  If the same plate is replugged 
+(different fiber mapping), and re-observed and processed, 
+a separate set of spectra are extracted for the new plate-night combination.
 
@@
 guiding has settled, the first science exposure is started.  The typical 
-exposure time is 900 seconds, but exposure times from 600 to 1200 second are
-not unexpected based on conditions and time constraints.  A second exposure
-is started immediately after the first science exposure is complete and read
-out.
-A fast Linux workstation onsite copies the data from the data acquisition host,
+exposure time is 900 seconds, but exposure times from 600 to 1200 second are not
+unexpected based on conditions and time constraints.  A second exposure is 
+started immediately after the first science exposure is complete and read out.
+A workstation onsite copies the data from the data acquisition host,
 and automatically performs a quick spectral extraction with a subset of the 
-steps explained here.  The results are posted to sos.apo.nmsu.edu, and
-the SDSS observers can check the signal-to-noise ratio (S/N) 
-obtained in each of the 4 spectroscopic
-cameras in real-time.  A plate is considered to have met SDSS survey standards
-if all 4 cameras have a cumulative (S/N)$^2 > 15$ at 
-fiber magnitudes of 20.33 and 20.06 in SDSS g and i filters respectively.
+steps explained here.  The results are posted to a local website to allow the 
+SDSS telescope operators to check the signal-to-noise ratio (S/N) 
+obtained in each of the 4 spectroscopic cameras in real-time.  
+A plate is considered to have met SDSS survey standards
+if all 4 cameras have a cumulative (S/N)$^2 > 15$ (per native spectral pixel) 
+at fiber magnitudes of 20.33 and 20.06 in SDSS g and i filters respectively.  
+The on-the-mountain automated reductions maintained to a good degree the quality
+standards of the spectroscopic survey and reduced the number of times plates 
+were reobserved due to lack of signal.
+
+\section{PIPELINE ORGANIZATION}
+\label{spplan}
+
+The full two-dimensional survey pipeline works in a small number of fixed steps
+that are run??/invoked sequentially.  These steps are summarized here, and then described in more detail in the remaining sections of this manuscript.
+\begin{enumerate}
+\item{Categorize and group every CCD frame taken on a single MJD night}
+\end{enumerate} 
+Describe the organization of data, and the types of data 
+used/required by the pipeline.
 
@@
 
-sdssproc:
-
-1000 lines of IDL code...
-
-read opEcalib
-     op something else
-
-Read file header, require...
-mention spHdrFix here.
+The raw spectroscopic data from each of the four CCDs is stored as 16 bit images
+with 2128 columns and 2069 rows.  The active area is represented by 2048x2048 
+24$\u$m pixels and read out by two indepedent amplifiers for each CCD.  
+Readout time for an unbinned image is nominally (27 ms per row) ?? seconds.
+A single IDL script is responsible for converting the raw fits files into 
+1) A header consisting on keyword-value pairs, 
+2) a 2048x2048 floating point data image, and 
+3) a corresponding inverse variance image.
+
+The main steps of the image processing for every frame are described here.
+
+- Identify correct exposure number and camera ID based on header and filename.
+- Apply neccesary header fixes based on external support file (spHeaderFix), and also purge any NaNs from header values.
+- Identify type of exposure (types??) and expected quality (test, unknown, bad, acceptable, good, excellent). 
+- Locate and read supporting operation files, including CCD layout (opConfig), CCD amplifier gains (opEcalib), and CCD bad pixel masks (opBC).
+- Subtract amplifier bias value computed as mean in appropriate overscan region.
+- Multiply by inverse gain to convert from bias corrected ADU to electrons.
+- Find columns subject to bleeding (when a saturated pixel that affects others along the same column).
+- Create a corresponding image representing variance, which is the per pixel sum of the data image, the gain corrected read noise squared, and a term limiting the reported pixel signal to noise at 100.??footnote this is an extra contribution to the variance given as (0.01 * data value)^2??
+- Convert variance image to inverse variance, and set inverse variance to zero (giving those pixels zero weight downstream) for any pixels with set in masks of bad or saturated pixels
+- Optionally subtract a bias image (usually created offline with a combined set of zero second exposures), no effect on inverse variance
+- Optionally divide by a flat-field image which represents the relative pixel-to-pixel response.  Correct inverse variance by 1/flat-field value squared.  
+-
+
+It is flexible, for instance it can be used just to read the header, and will not read the entire raw frame.
+
+There were also many special case situations that are 
+Case 1:  shifted rows due to electronic issues on select nights in 2000.
+Case 2:  Correct image for data acquired on nights when images were readout with shutter open.
+
+??applycrosstalk, check svn blame??
 
@@
 
-trace320, Trace crude, and trace fitting
+Top level, call frm spcalib as of r54990 (20020301):  
+ - Trace320crude, then fit with 5th-order polynomial.
+ - The current trunk code does not match what was used for the SDSS legacy reductions, especially trace_cen.  What revision corresponds to the one used in the DR reductions??
+
+ - Find initial fiber positions on each flat-field frame acquired during the 
+plate observation sequence.  The final structure is a 320x2048 set of column 
+positions representing the best estimate of the fiber column position in each 
+row of the frame.
+ - Given the camera optics the column positions bow inward towards the middle 
+of the CCD frame.  The fibers are more closely spaced in column position for 
+rows near the middle of the CCD.  The set of 21 rows about the center of the 
+CCD is chosen to find the initial fiber positions.  A median\footnote{iterative
+median, see appendix} is performed at each column positions, and a single row
+is analyzed representing the median value per column.  The "median" central row
+is cross correlated with a template representing expected fiber positions taking
+into account fiber spacings and bundle separations.
+- Given the best correlation offset (and minimum correlation value of ??), 
+tweak each expected position from the template to the best empirical position 
+in the median row by calculating a flux-weighted mean over a window of 6 pixels.
+
+to locate the majority of profile peaks (see Figure ??? for examples of fiber
+profiles in a typical SDSS frame).
+
+Find approximate centers of all 320 fiber profiles for a single row, and note uncertain identifications due to bad pixels or falling off CCD frame.
+The profile position is calculated as the interpolated peak in cross-sectional counts.
+Cross correlate the identified positions with a template based on known fiber spacings.
+??  When did trace_cen move to parinfo standard ??
+
+From initial centers move up and down the CCD images and "trace" the position centers by calculating the best count-weighted center over a fixed window based on the previous center.
+
+Then perform a polynomial fit on all 320 fibers profiles as column positions versus row number.
+A 4th order chebyshev??
+polynomial is sufficient.
+The polynomial fit is purely empirical and there is no optical model or other physical constraint on the polynomial solutions.
+
+Next steps??
 
@@
 
+\begin{table}
+\begin{tabular}{ccc}
+\hline
+DR &  2d version & rerun ? \\
+\hline
+\hline
+EDR - DR4 & 4\_?? & \\
+DR5 & 5\_?? & \\
+DR6 & ?? & 25 \\
+DR7 & ?? & \\
+\hline
+\end{tabular}
+\end{table}
+
 \section{Testing velocities}
@@
 and choosing the best-fit Elodie star using $\chi^2$ minimization.  (ref???)
-The derived veocites have a median offset of $-5.5\kms$, with a sigma-clipped
+The derived velocites have a median offset of $-5.5\kms$, with a sigma-clipped
 dispersion of $3.0 \kms$ (see Figure \ref{fig_m67}).
@@
 % Fiber 301 at RA= 132.01817 Dec= 12.167990 v(CAT)= 12.10 v(Elodie)= 104.35960
+% What's the third one?  SMB 20110302
 
@@
 SDSS spectroscopic
-plates observed between MJD 2451600 to 2453600.  Each plate is
+plates observed between MJD 51600 to 53600.  Each plate is
 represented by 3 symbols showing the median RMS guiding residuals

### 2011-03-17 doc/paper/paper2d.tex
Further comments and spatial tracing

@@
 Top level, call frm spcalib as of r54990 (20020301):  
- - Trace320crude, then fit with 5th-order polynomial.
- - The current trunk code does not match what was used for the SDSS legacy reductions, especially trace_cen.  What revision corresponds to the one used in the DR reductions??
-
- - Find initial fiber positions on each flat-field frame acquired during the 
-plate observation sequence.  The final structure is a 320x2048 set of column 
-positions representing the best estimate of the fiber column position in each 
-row of the frame.
- - Given the camera optics the column positions bow inward towards the middle 
-of the CCD frame.  The fibers are more closely spaced in column position for 
-rows near the middle of the CCD.  The set of 21 rows about the center of the 
-CCD is chosen to find the initial fiber positions.  A median\footnote{iterative
-median, see appendix} is performed at each column positions, and a single row
-is analyzed representing the median value per column.  The "median" central row
-is cross correlated with a template representing expected fiber positions taking
-into account fiber spacings and bundle separations.
-- Given the best correlation offset (and minimum correlation value of ??), 
-tweak each expected position from the template to the best empirical position 
-in the median row by calculating a flux-weighted mean over a window of 6 pixels.
-
-to locate the majority of profile peaks (see Figure ??? for examples of fiber
-profiles in a typical SDSS frame).
-
-Find approximate centers of all 320 fiber profiles for a single row, and note uncertain identifications due to bad pixels or falling off CCD frame.
-The profile position is calculated as the interpolated peak in cross-sectional counts.
-Cross correlate the identified positions with a template based on known fiber spacings.
-??  When did trace_cen move to parinfo standard ??
-
-From initial centers move up and down the CCD images and "trace" the position centers by calculating the best count-weighted center over a fixed window based on the previous center.
-
-Then perform a polynomial fit on all 320 fibers profiles as column positions versus row number.
-A 4th order chebyshev??
-polynomial is sufficient.
-The polynomial fit is purely empirical and there is no optical model or other physical constraint on the polynomial solutions.
-
+ - Trace320crude, then fit with 7th-order polynomial for column positions versus CCD row.
+ - First, find all 16 bundles of 20 fibers by assuming known fiber (6.25 pixel)
+and bundle (8 pixel) spacings.
+Locate peaks by convolving the composite set of 21 central rows with a peak finding kernel\footnote{The kernel used was [-1, -1, 1, 2, 1, -1, -1]} that must match the known fiber pattern.
+??please make more readble??
+ - Starting with the initially determined column locations of all 320 fibers, 
+improve estimate of centroids using a simple flux-weighted mean over a window of 6 pixels centered at the current position.  
+ - Step in rows sequentially up and down using the previous row's positions as the initial window position in the flux-weighted mean.
+ - Mask fibers which had problematic centroids, and replace centroids for all rows with interpolated values from neighboring traces.
+ - Fit each spatial trace with a 7th order polynomial, with iteration and rejecting pixels which fall outside of 0.15 pixels from the best fit.
 Next steps??
@@
 \hline
-EDR - DR4 & 4\_?? & \\
-DR5 & 5\_?? & \\
-DR6 & ?? & 25 \\
-DR7 & ?? & \\
+DR2 & 4\_10\_4  & 23 \\
+DR3 & 4\_10\_4  & 23 \\
+DR4 & 4\_10\_4  & 23 \\
+DR5 & 4\_10\_4 & 23 \\
+DR6 & 5\_3\_3  & 25 \\
+DR7 & 5\_3\_12 & 26 \\
+DR8 & 5\_3\_12 & 26 \\
 \hline
@@
 
+?? Refer to DR5? paper and Tremonti's SDSS paper which compares 
+spectrophotometric colors to photo colors ??
+
 %------------------------------------------------------------------------------
 \subsection{GALACTIC REDDENING}
+
+?? Drop galactic reddening, users should known when or when not to correct for it ??
 
@@
 
+\bibitem[Subbarao et al. (2002)]{2002SPIE.4847..452S} SubbaRao, M., et al.\ 2002, SPIE, 4847, 452
+
 \bibitem[Adelman-McCarthy et al.(2006)]{2006ApJS..162...38A} 
 Adelman-McCarthy, J.~K., et al.\ 2006, \apjs, 162, 38
+
+\bibitem[Adelman-McCarthy et al.(2007)]{2007ApJS..172..634A} 
+Adelman-McCarthy, J.~K., et al.\ 2007, \apjs, 172, 634
+
+\bibitem[Abazajian et al.(2009)]{2009ApJS..182..543A} Abazajian, K., et 
+al.\ 2009, \apjs, 182, 543
 
@@
 
+?? Alot of this is already published in the DR papers or on the DR websites ??
 This spCFrame files contain the extracted spectra for a single CCD of a
@@
 
+?? Just refer to EDR paper, or one of the DR papers, and drop this section ??
+
 The target selection sets bits in two flags called PRIMTARGET and
@@
 is in the center of the beam in this Schmidt camera design.
+?? Drop this figure and refer to the Smee spectrograph paper ??
 \label{fig2}
@@
 %\plotone{plot_bestthru.ps}
-\figcaption{ Throughput of the SDSS spectroscopic system,
+\figcaption{ 
+?? A similar figure is shown in the EDR paper, let's save space and refer to that??
+Throughput of the SDSS spectroscopic system,
 as measured relative to the photon density expected above

### 2012-05-14 doc/paper/paper2d.tex
Correct typo as pointed out by Ning-Chen Sun.

@@
 keeping the same number of breakpoints, but a quadratic polynomial fit is
-allowed as a function of wavelength.  This triples the number of free 
+allowed as a function of fiber number.  This triples the number of free 
 parameters in the linear least-squares fit, but the values are still 

### 2012-06-19 doc/paper/paper2d.tex
Update.

@@
 
-\title{The Sloan Digital Sky Survey: Two-dimensional Spectroscopic Reduction Pipeline}
+\title{The Sloan Digital Sky Survey Two-dimensional Spectroscopic Reduction Pipeline}
 
 \author{
-Scott Burles\altaffilmark{\ref{email}}
-and David J. Schlegel\altaffilmark{\ref{Princeton}}
+David J. Schlegel\altaffilmark{\ref{LBL}}
+and Scott Burles\altaffilmark{\ref{email}}
 }
 \altaffiltext{1}{sburles@gmail.com \label{sburles}}
-\altaffiltext{2}{Princeton University Observatory, Peyton Hall, Princeton,
-NJ 08544 \label{Princeton}}
+\altaffiltext{2}{Lawrence Berkeley National Laboratory,
+1 Cycltotron Road, Berkeley, CA 94720 \label{LBL}}
 
@@
 \begin{abstract}
-The Sloan Digital Sky Survey (SDSS) fiber-fed spectrographs have been used
-to obtain spectra for ~$600,000$ galaxies, quasars and stars
-with a resolution of ~2000 over the wavelength range 3800\AA -9600\AA.
-We present the pipeline design and algorithms for reducing the raw data to
-wavelength-calibrated, flux-calibrated one-dimensional spectra.
-It is designed to automatically reduce survey quality data sets and output 
-uniformly processed results.  No human intervention is required in any step 
-of the pipeline beyond invoking the top level command pointing to the 
-applicable data.
-The wavelength calibration is shown to have an accuracy of $3 \kms$,
-and the broad-band spectrophotometry to have an accuracy of better than $6\%$.
-What about reported errors and their accuracy/robustness?
-The reported errors capture the statistical and the expected have
+We describe the data reduction pipeline for the spectroscopic data
+reduction for the optical spectrographs on the Sloan Telescope.
+These spectrographs were used to obtain spectra of 900,000 galaxies,
+100,000 QSOs and 300,000 stars during the Sloan Digital Sky Survey
+(SDSS) and Sloan Digital Sky Survey II (SDSS-II) from 1999 through 2009.
+The instrument was rebuilt in 2009 for the Sloan Digital Sky Survey III
+(SDSS-III) and will be used to complete a survey of an additional
+1.35 million galaxies and 200,000 QSOs by 2014.
+We present the pipeline design and algorithms for reducing the raw
+data to wavelength-calibrated, flux-calibrated one-dimensional spectra
+and per-pixel error estimates.
+The pipeline has been used to automatically reduce the survey data
+sets and output uniformly-processed results.
+The wavelength calibration is demonstrated to have an accuracy
+of $3\kms$ and the broad-band spectrophotometry to have an accuracy
+of $6\%$ for SDSS-I and II and $9\%$ for SDSS-III.
+These spectra are of sufficient quality to allow automated redshift
+fits to $99\%$ of the galaxies in SDSS-I and II and $\98\%$ of the much
+fainter galaxies in SDSS-III, and catestrophic redshift errors
+of much less than $1\%$ in all cases.
 
@@
 \label{sec_intro}
-The Sloan Digital Sky Survey (SDSS) is a 5-band photometric survey of
-approximately $10,000$ square degrees of the Northern sky and a concurrent
-redshift survey of up to a million galaxies and 100,000 quasars
-selected from the imaging survey (York et al 2000).  The primary 
-science goals of the project are to provide the data to investigate the
-large scale structure of the Universe as well as other extragalactic and 
-galactic science.
+The Sloan Telescope was designed to provide 5-band imaging
+of approximately one-quarter of the sky and spectroscopic follow-up
+of millions of galaxies, QSOs and stars in that same footprint.
+The original SDSS operated from 1998-2005 and
+imaged 7600 deg$^2$ with spectroscopic follow-up of ???.
+SDSS-II operated from 2005-2008 as a combination of three programs:
+spectroscopic follow-up in the remainder of the original SDSS footprint,
+a supernova search from repeated imaging of an equatorial stripe,
+and imaging and spectroscopic follow-up for Galactic stars (SEGUE).
+
+The SDSS spectroscopic pipeline (idlspec2d) was written in support
+of the original galaxy and QSO redshift survey.
+This pipeline was modified to provide more accurate redshifts
+and spectrophotometry in support of the SEGUE program in SDSS-II.
+The pipeline was again modified to support the rebuilt spectrographs
+for the Baryon Oscillation Spectroscopic Survey (BOSS) in SDSS-III,
+which is observing targets 2.5 mag fainter than SDSS.
+Finally, many of the algorithms and code have been used for
+the spectroscopic reductions of the DEEP2 survey.
+
+The pipeline is written in the IDL language with some C bindings.
+It makes use of the Goddard IDL astronomy library.
+The code and all previous versions are publicly-available
+through the SDSS-III project at {sdss3.org}.
+
+The goal of the SDSS spectroscopic pipeline is to produce the
+best estimate of spectral intensity as a function of wavelength.
 
@@
 \hline
-DR &  2d version & rerun ? \\
+Data Release & Instrument & Code version ? \\
 \hline
 \hline
-DR2 & 4\_10\_4  & 23 \\
-DR3 & 4\_10\_4  & 23 \\
-DR4 & 4\_10\_4  & 23 \\
-DR5 & 4\_10\_4 & 23 \\
-DR6 & 5\_3\_3  & 25 \\
-DR7 & 5\_3\_12 & 26 \\
-DR8 & 5\_3\_12 & 26 \\
+DR1 & SDSS & ?   \\
+DR2 & SDSS & v4\_10\_4   \\
+DR3 & SDSS & v4\_10\_4   \\
+DR4 & SDSS & v4\_10\_4   \\
+DR5 & SDSS & v4\_10\_4  \\
+DR6 & SDSS & v5\_3\_3   \\
+DR7 & SDSS & v5\_3\_12  \\
+DR8 & SDSS & v5\_3\_12  \\
+DR9 & BOSS & ? \\
 \hline

### 2015-12-04 doc/algorithms/algorithms.tex
Adding stub for documentation about algorithm dataflow

\title{idlspec2d dataflow}
\author{Stephen Bailey, Julian Bautista, David Schlegel}
\date{\today}

\documentclass[12pt]{article}

\usepackage{amsmath}

\begin{document}
\maketitle

\abstract{
This note documents the idlspec2d algorithm dataflow.
The majority of the code was written by David Schlegel;
the purpose of this document is to succinctly describe each of the steps.
}

%-------------------------------------------------------------------------
\section{Overview}

BOSS has 2 spectrographs with 500 fibers each, grouped in 25 bundles of 20
fibers each.  The light is split into a blue channel and a red channel, for
a total of 4 CCD images per exposure. The CCD $y$ coordinate is the spectral
dispersion direction (larger $y$ is larger $\lambda$) and
larger $x$ is larger fiber number, though the spectral ``traces'' in
$y$~{\it vs.}~$x$ are curved and do not exactly align with CCD columns.

{\it Raw electrons} are extracted from the CCD images using row-by-row extractions similar to Horne~1986 by fitting Gaussians plus a polynomial
background to each CCD row for each bundle of 20 fibers.

{\it Fiber flats} correct for fiber-to-fiber variations by comparing the
differences between fibers equally illuminated by a smooth flat lamp spectrum.

{\it Sky model} is derived from flat fielded electrons of sky fibers and then
interpolated to the locations of every science fiber and subtracted.

{\it Flux calibration} vectors model the instrument and atmospheric throughput
per exposure by comparing standard star spectra to a set of models of known
flux.

{\it Flux correction} vectors adjust for flux mis-calibrations with low-order
polynomials per-fiber per-exposure to make different exposures of the same object consistent with each other.

{\it Flux distortion} vectors model variations in the throughput
across the focal plane.

Putting these together --- \newline
``flat fielded sky subtracted electrons'' in spFrame:
\begin{equation}
F_e = \mathrm{electrons} / (\mathrm{superflat} \cdot \mathrm{fiberflat}) -
    \mathrm{skymodel}
\end{equation}
become the ``calibrated flux'' in spCFrame:
\begin{equation}
F = (F_e / \mathrm{calib}) \cdot
    \mathrm{fluxcorr} \cdot \mathrm{fluxdistort}
\end{equation}

The following sections describe each of these steps in detail.

%-------------------------------------------------------------------------
\section{Extracted electrons to calibrated flux}

\subsection{Extraction}

\subsection{Fiber flats and trace locations}

\subsection{Wavelength solution and PSF shape}

\subsection{Sky subtraction}

\subsection{Flux calibration}

\subsection{Flux correction}

\subsection{Flux distortion}

The {\it flux distortion} vectors are parameterized in terms of magnitude
(i.e.~log-flux) that are achromatic with $x$, $y$, $x^2$, $y^2$, $xy$,
where those are linear coordinates XFOCAL,YFOCAL from the plugmap.
There are also chromatic terms that scale as
$\tilde \lambda = 1-(5070/\lambda)^2$,
since that function gives an equal effect between 3900 and 5070~\AA\
as between 5070 ang 9000~\AA.
There are also magnitude offsets as a function of spectrograph ID,
and a chromatic offset as a function of spectrograph ID.  The 14 parameters are:
\begin{equation}
\begin{split}
    F_\mathrm{new} &= F_\mathrm{orig}
        \left(1 + a_0 s_1 + a_1 s_2 \right) \mathrm{exp} \left( \right. \\
&            a_2 x + a_3 y + a_4 x y + a_5 x^2 + a_6 y^2 + \\
&            a_7 \tilde \lambda x + a_8 \tilde \lambda y + \\
&             a_9 \tilde \lambda s_1 + a_{10} \tilde \lambda s_2 +
             a_{11} \tilde \lambda^2 s_1 + a_{12} \tilde \lambda^2 s_2
        \left. \right)
\end{split}
\end{equation}
where $s_1$=1 if specid=1 else 0 and $s_2$=1 if specid=2 else 0.

%-------------------------------------------------------------------------
\section{Coaddition}

%-------------------------------------------------------------------------
\section{Code}

\begin{tabular}{lp{30mm}p{30mm}}
Code & Inputs & Outputs \\
\hline
spreduce2d  & spPlan2d  & \\
\hline
  spreduce    & sdR \newline plPlugMapM & \\
\hline
    spcalib     &       & spArc \newline spFlat \newline spFrame \\
\end{tabular}

\end{document}

### 2016-06-02 doc/algorithms/algorithms.tex
adding stuff to fluxdistortion
adding fluxcorr description

@@
 
+Individual exposures are initially flux calibrated with no constraint that the same object has the same flux across different exposures.
+Emperical ``fluxcorr'' vectors are broadband corrections to bring the different exposures into alignment for each object prior to coaddition.
+In DR13 and prior, these were implemented for each spectrum by minimizing
+\begin{equation}
+    \chi^2_i = \sum_\lambda {
+        (f_{i\lambda} - f_{\mathrm{ref},\lambda} / a_{i\lambda} )^2 \over
+        (\sigma_{i\lambda}^2 - \sigma_{\mathrm{ref},\lambda}^2 / a_{i\lambda}^2)
+        }
+\end{equation}
+where $f_{i\lambda}$ is the flux of exposure $i$ at wavelength $\lambda$;
+$f_{\mathrm{ref},\lambda}$ is the flux of the selected reference exposure;
+and $a_{i\lambda}$ are low order Legendre polynomials.
+%--- If you want more detail:
+ The number of polynomial terms is dynamic, up to
+ a maximum of 5 terms.  Higher order terms are added only if they improve
+ the $\chi^2$ by 5 compared to one less term.
+This approach is biased toward small $a_{i\lambda}$ since that inflates the denominator to reduce the $\chi^2$.
+
+For DR14, we solve the fluxcorr vectors relative to a common weighted coadd $F_\lambda$ which is treated as noiseless compared to the individual exposures.
+\begin{equation}
+    \chi^2_i = \sum_\lambda {
+        (f_{i\lambda} - F_\lambda / a_{i\lambda})^2 \over
+        \sigma_{i\lambda}^2
+        }
+\end{equation}
+We additionally include an emperically tuned prior that $a_{i\lambda} \sim 1$ to avoid large excursions in the solution for very low signal-to-noise data.
+
+ The DR14 fluxcorr terms are Chebychev polynomials instead of Legendre for no
+ good reason.  We actually solve for ((f - F a)/sigma)^ and then return 1/a .
+ The prior is weighted by the data weights such that the relative strength of
+ the prior vs. the data is approximately independent of S/N.  I'm not convinced
+ that is the best thing to do but it worked better than having a fixed
+ strength prior.
+
 \subsection{Flux distortion}
@@
 There are also magnitude offsets as a function of spectrograph ID,
-and a chromatic offset as a function of spectrograph ID.  The 14 parameters are:
+and a chromatic offset as a function of spectrograph ID.  The 13 parameters are:
 \begin{equation}
@@
 where $s_1$=1 if specid=1 else 0 and $s_2$=1 if specid=2 else 0.
+
+Only \textsc{SPECTROPHOTO\_STD} or \textsc{REDDEN\_STD} objects are used to compute the flux distortion.
+The procedure minimizes differences between spectro-flux and photometry (CALIBFLUX).
+For now, only g, r and i-bands are used.
 

### 2016-06-04 bin/fluxcorr_prior.py
correcting small bugs in fluxcorr_prior.py

@@ -205,6 +205,9 @@ plandir = os.path.dirname(os.path.abspath(planfile))
 framefiles = read_plan(planfile)
 for camera in ('b1', 'b2', 'r1', 'r2'):
     flux, ivar, goodframes = read_data(plandir, framefiles[camera], xythrucorr=xythrucorr)
+    if len(goodframes)==0:
+        print 'No goodframes found'
+        continue
     fluxcorr = calc_fluxcorr(flux, ivar, prior=1.0)
     addterm = N.zeros(fluxcorr[0].shape)
    
